```{r setup_0605, include=FALSE}
library(tidyverse)
library(gt)
```

# Part 5: Generalised Linear Models - Poisson {#sec-p5_glms_binomial}

::: {.callout-note .partmenu #parts-0605}
## Sections
- @sec-fitting_pois_glms
- @sec-summary_0605
:::

:::{.callout-tip .objectives}
#### Learning objectives
By the end of this part of the practical you will be able to:

:::

A binomial GLM is appropriate when the response variable represents one of two possible outcomes. A Poisson GLM can be used to model count data when a Poisson distribution (@sec-comparing_sim_real) is appropriate. In a Poisson GLM, we model how $\lambda$, the expected mean count, changes with our explanatory variables.

For example, with our Poisson data, let's test the following hypotheses:

**Null hypothesis**: There is no association between the number of non-Apidae insect species and the expected number of Apidae species found at a site.

**Alternative hypothesis**: There is an association between the number of non-Apidae insect species and the expected number of Apidae species found at a site.

## Fitting Poisson GLMs in R {#sec-fitting_pois_glms}

First we need to read our data, and add a column containing the number of non-Apidae species at each site.

```{r pois_glm_apidae_data, message=FALSE}
site_metadata <- read_csv("data/site_metadata.csv")
site_metadata <- site_metadata |>
  mutate(n_non_apidae = n_unique - n_apidae)
```

Now we can apply the GLM, with `family = poisson` for the Poisson model.

```{r pois_glm_apidae}
glm_apidae <- glm(n_apidae ~ n_non_apidae,
                  family = poisson,
                  data = site_metadata)
summary(glm_apidae)
```

The estimated coefficient for the number of non-Apidae species is `r round(coef(glm_apidae)["n_non_apidae"], 3)`. This coefficient is positive, indicating that sites with more non-Apidae insect species are expected to also contain more Apidae species.

The Poisson GLM uses a **log link**, so the coefficients represent changes in the log of the expected count. We can exponentiate the coefficient to make it easier to interpret:

```{r exponentiate_pois_glm_apidae}
print (exp(coef(glm_apidae)["n_non_apidae"]))
```

For each additional non-Apidae species, the expected number of Apidae species is therefore multiplied by `r round(exp(coef(glm_apidae)["n_non_apidae"]), 3)`, corresponding to an increase of approximately `r round((exp(coef(glm_apidae)["n_non_apidae"]) - 1) * 100, 1)`%.

The p-value associated with this coefficient is `r format.pval(coef(summary(glm_apidae))["n_non_apidae", "Pr(>|z|)"], digits = 3)`. We therefore reject the null hypothesis and conclude that there is evidence of an association between the number of non-Apidae insect species and the expected number of Apidae species found at a site.

Let's plot this relationship.

```{r plot_glm_apidae}
ggplot(site_metadata,
       aes(x = n_non_apidae, y = n_apidae)) +
  geom_point() +
  geom_smooth(
    method = "glm",
    method.args = list(family = poisson)
  ) +
  theme_bw()
```

## Review {#sec-summary_0605}

[↑ top](#)

::: {.callout-note #note-summary_0605}

### Summary


:::

For this part of the practical, the main files you have created are:

These are the files you should select when making your Git commit.


Remember to save and commit your work before moving on to Part 6.

{{< include ../includes/git_commit.qmd >}}
