---
format: html
practical: Practical_6_Further_Statistics
doc-id: "0601"
author: "Katy Brown"
---


```{r setup_0601, include=FALSE}
library(tidyverse)
library(gt)
```

# Part 1: Functions {#sec-p1_functions}
::: {.callout-note .partmenu #parts-0601}
## Sections

- @sec-writing_functions
- @sec-summary_0601
:::

:::{.callout-tip .objectives #objectives-0601}

#### Learning objectives
By the end of this part of the practical you will be able to:

* Understand what functions are and when they are used. 
* Write functions to automate aspects of your analysis.
* Use a function to run a multi-step analysis on a data frame.

:::

In this practical, you will need one Quarto documents:

* `prac6_exercises.qmd` for basic exercises

## Writing functions {#sec-writing_functions}

In the practical course so far, we have used many of the different **functions** (see @sec-functions) which are built into R and provided by additional packages.

However, it is also possible to **write your own functions**. This means you can use the same set of instructions multiple times, possibly with different inputs.

Functions have several advantages:

* Using the same code again saves time.
* If you want to change something in the code, you only have to change it once.
* Because you are writing less code, you are less likely to make mistakes.

Here is a simple example of a function, to calculate the equivalent age of a dog if it were a human, its age in *dog years*. According to [the BBC](https://www.bbc.co.uk/news/magazine-22479412), for a medium dog the calculation is:

$$
d = 21 + \left(4 \times (\text{h} - 2)\right)
$$

Where *d* is the age in dog years and *h* is age in human years.

We can write a function to calculate this. We define a function using the function `function()`. It is best to give the function a descriptive name.

The steps of the function are then written in curly brackets (`{}`) on the following lines.

The `return` statement defines what the **output** of the function will be.

```{r first_function}
human_years_to_dog_years <- function(age_in_human_years) {
  age_in_dog_years <- 21 + (4 * (age_in_human_years - 2))
  return(age_in_dog_years)
}
```

So here we are saying:

* <code>[human_years_to_dog_years <- function(age_in_human_years)]{.orange-text}</code>: I am defining a function, it is called `human_years_to_dog_years` and takes one argument, `age_in_human_years`.
* <code>[{]{.orange-text}</code>: Here are the steps.
* <code>[age_in_dog_years <- 21 + (4 * (age_in_human_years - 2))]{.orange-text}</code>: Run this equation on the variable `age_in_human_years` and store the result as the variable `age_in_dog_years`.
* <code>[return (age_in_dog_years)]{.orange-text}</code>: Return the value of the variable `age_in_dog_years` as the output of the function
* <code>[}]{.orange-text}</code>: That's all.

Let's use our function to calculate the age in dog years of some dogs.

Monty is 6 in human years.

```{r dog_age_1}
print(human_years_to_dog_years(6))
```


Noodle is 2 in human years:

```{r dog_age_2}
print(human_years_to_dog_years(2))
```

We can also apply our function to a column in a data frame.

Let's say we have this data frame of dogs.

```{r dog_tibble}
dog_data <- tibble(
  Name = c(
    "Biscuit",
    "Snowball",
    "Pickle",
    "Artemis",
    "Bert"
  ),
  Age_Human_Years = c(6, 1, 12, 4, 15)
)
gt(dog_data)
```

Using our `human_years_to_dog_years()` function, we can calculate the age in dog years of each dog.

[comment]: todo_katy link to the mutate explanation once I've written it.

::: {.callout-note #note-mutate}

Remember that the `mutate()` function allows us to create a new column in a data frame by applying a function to an existing column.
:::

```{r dog_years_df}
dog_data_updated <- dog_data |>
  mutate(Age_Dog_Years = human_years_to_dog_years(Age_Human_Years))

gt(dog_data_updated)
```


There are two important things to learn about functions:

**1. In order to run the function, you need to call it**

So, if I just run this code:

```{r function_call_1}
human_years_to_dog_years <- function(age_in_human_years) {
  age_in_dog_years <- 21 + (4 * (age_in_human_years - 2))
  return(age_in_dog_years)
}
```

R now knows this function exists, but it hasn't been used yet.

If we want to actually run it, we need to add a **function call**.

```{r function_call_2}
human_years_to_dog_years <- function(age_in_human_years) {
  age_in_dog_years <- 21 + (4 * (age_in_human_years - 2))
  return(age_in_dog_years)
}
# function call
print(human_years_to_dog_years(10))
```


**2. Variables you create inside a function only exist inside the function.**

So, the following will work - inside the function, between the curly brackets, we can access the variable `age_in_dog_years`.

```{r function_vars1}
human_years_to_dog_years <- function(age_in_human_years) {
  print(paste("Age in human years:", age_in_human_years))
  age_in_dog_years <- 21 + (4 * (age_in_human_years - 2))
  return(age_in_dog_years)
}
# function call
print(paste("Age in dog years:", human_years_to_dog_years(10)))
```

But this will not work, because outside of the function, the variable `age_in_dog_years` does not exist.

```{r function_vars2, error=TRUE}
human_years_to_dog_years <- function(age_in_human_years) {
  age_in_dog_years <- 21 + (4 * (age_in_human_years - 2))
  return(age_in_dog_years)
}
result <- human_years_to_dog_years(10)
print(paste("Age in human years:", age_in_dog_years))
```

:::{.callout-exercise #ex-dog_function}

In your `prac6_exercises.qmd` Quarto document, write a function to convert dog years to human years (the reverse of the above). Round the answer to the nearest whole number.

Apply your function to the following:

* Billy, a human aged 180 dog years
* Ahmed, a human aged 44 dog years
* The `Age_Dog_Years` column of the following data frame of four humans:

```{r human_tibble}
human_data <- tibble(
  Name = c(
    "Aisha",
    "Mateo",
    "Priya",
    "Daniel"
  ),
  Age_Dog_Years = c(76, 222, 239, 123)
)
gt(human_data)
```

::: {.callout-note .copycode #copy-libs_tcga}
### Include this code

Include this code at the top of your `prac6_exercises.qmd` Quarto file.

```{r libraries_tcga}
library(tidyverse)
library(gt)
```
:::

::: {.callout-answer collapse="true" #dog_function_ans}
```{r ex-dog_function_ans1}
dog_years_to_human_years <- function(age_in_dog_years) {
  age_in_human_years <- ((age_in_dog_years - 21) / 4) + 2
  return(round(age_in_human_years))
}

print(paste(
  "Billy is", dog_years_to_human_years(180),
  "human years old"
))
print(paste(
  "Ahmed is", dog_years_to_human_years(44),
  "human years old"
))
```

```{r ex-dog_function_ans2}
human_data_updated <- human_data |>
  mutate(Age_Human_Years = dog_years_to_human_years(Age_Dog_Years))

gt(human_data_updated)
```
:::
:::
## Review {#sec-summary_0601}

[↑ top](#)

::: {.callout-note #note-summary_0601}

### Summary


:::

For this part of the practical, the main files you have created are:

These are the files you should select when making your Git commit.


Remember to save and commit your work before moving on to Part 2.

{{< include ../includes/git_commit.qmd >}}
