---
format: html
practical: Practical_4_Statistics
doc-id: "0401"
author: "Katy Brown"
---

```{r setup_0401, include=FALSE}
library(tidyverse)
library(gt)
library(osbng)
library(sf)
library(maps)
insect_metadata <- read_csv("data/insect_metadata.csv")
```

# Part 1: UK Pollinator Monitoring Scheme Dataset {#sec-p1_data}

::: {.callout-note .partmenu #parts-0401}
## Sections

- @sec-dataset_pollinators
- @sec-summary_0401
:::

:::{.callout-tip .objectives}
#### Learning objectives
By the end of this part of the practical you will:

* Understand the structure of the pollinator dataset
* Have recapped some data frame operations from @sec-p1_basic_data_frame_operations and @sec-p2_complex_data_frame_operations.

:::

## Sync your Git repository
{{< include ../includes/git_pull.qmd >}}

For this practical, we recommend that you create two Quarto documents:

* `{{< meta practical >}}/pollinators.qmd` - analysis of the pollinator dataset
* `{{< meta practical >}}/prac4_exercises.qmd` - general practice exercises

We will use the following research dataset repeatedly throughout both this practical and Practical 6 (Further Statistics) and visit others as needed.

## Pollinator dataset {#sec-dataset_pollinators .core}

[**Core**]{.core-text}

The main dataset we will use in this practical is [pan-trap survey data](https://catalogue.ceh.ac.uk/documents/9df83130-e98f-48b0-bbfc-32f5920eedb0) from the UK Pollinator Monitoring Scheme (POMS). In this scheme, multiple [pan-traps](https://en.wikipedia.org/wiki/Pan_trap) are set in each of a set of specific 1km squares across the UK. On four days each summer, volunteers will set out traps for six hours, then send the insects to be counted. Flowering plants within a 2m radius of the trap are also recorded.

[comment]: todo_katy link to the preprocessing.

The data from the scheme is provided as data per trap, for this practical we have amalgamated the data across all dates and across each 1km square. The preprocessing of the data is described in XXXX. 

You have been provided with three data files:

* In the **insect counts** CSV file - `data/insect_counts.csv` each row represents a combination of an insect species and a sampling site, and the total number of times that insect was observed at that site. Many of the counts will be zero.

* In the **flower counts** CSV file - `data/flower_counts.csv` each row represents a combination of a flower species and a sampling site, plus TRUE or FALSE depending if that flower was ever observed within 2m of the site.

* In the **site metadata** CSV file - `data/site_metadata.csv`, each row represents a single 1km square sampling site. The data frame contains information about the geographical location, plus land use and climate at the sample site. Additional climate data has been added based on [WorldClim](https://nordata.physics.utoronto.ca/en/annexes-datasets-descriptions/72-precipitation/worldclim-version-2/) data, which has detailed climate data by square averaged from 1970-2000. Various additional statistics have been calculated based on the information in the original data frame.

<details>
<summary>Data citation</summary>

UK Pollinator Monitoring Scheme (2024). 1km square pan trap data from the UK Pollinator Monitoring Scheme, 2017-2021. NERC EDS Environmental Information Data Centre. (Dataset).

The UK Pollinator Monitoring Scheme (PoMS) is a partnership funded jointly by UKCEH and JNCC (through funding from Defra, Scottish Government, Welsh Government, and DAERA). UKCEH’s contribution is funded by the Natural Environment Research Council award number NE/R016429/1 as part of the UK-SCAPE programme delivering National Capability. Between 2017 and 2021, PoMS was funded by UKCEH and Defra (England), Welsh Government, Scottish Government, DAERA (Northern Ireland), and JNCC.
PoMS is indebted to the many volunteers who carry out surveys and contribute data to the scheme.

</details>

::: {.callout-exercise #ex_summarise_bees .rec}

### Revise data frame operations

[**Recommended**]{.rec-text}

In your `pollinators.qmd` Quarto file:

* Read the three CSV files:

  * `data/insect_counts.csv`
  * `data/flower_counts.csv`
  * `data/site_metadata.csv`

* Count how many unique sites there are. Site ID is in the `x1km_square` column of the insect counts data frame and site metadata data frame. 
* Count how many unique observations of insects there are - including multiple observations of the same insect at the same site. Number of observations is in the `n_observed` column of the insect counts data frame.
* Find the insect species which was observed most frequently and how many times it was observed. Insect species is in the column `species` of the insect counts data frame.

::: {.callout-answer #ex_summarise_bees_ans}

[Read the three CSV files]{.blue-text}

```{r summarise_bees_ans1, message=FALSE}
insect_counts <- read_csv("data/insect_counts.csv")
flower_counts <- read_csv("data/flower_counts.csv")
site_metadata <- read_csv("data/site_metadata.csv")
```

[Count how many unique sites there are. Site ID is in the `x1km_square` column of the insect counts data frame and site metadata data frame. ]{.blue-text}

```{r summarise_bees_ans2, message=FALSE}
site_count <- nrow(site_metadata)
print(site_count)
```
There are `r site_count` sites.

[Count how many unique observations of insects there are - including multiple observations of the same insect at the same site. Number of observations is in the `n_observed` column os the insect counts data frame]{.blue-text}

```{r summarise_bees_ans3, message=FALSE}
total_insect_obs <- insect_counts |>
  pull(n_observed) |>
  sum()
print(total_insect_obs)
```
There are `r format(total_insect_obs, scientific=FALSE)` unique insect observations.

[Find the insect species which was observed most frequently and how many times it was observed. Insect species is in the column `species` of the insect counts data frame and the insect metadata data frame.]{.blue-text}

```{r summarise_bees_ans4, message=FALSE}
insect_totals <- insect_counts |>
  group_by(species) |>
  summarise(n_obs = sum(n_observed))

insect_totals <- insect_totals |>
  arrange(desc(n_obs))

gt(head(insect_totals))
```

The most commonly observed species was `r pull(insect_totals, species)[1]`, which was observed `r pull(insect_totals, n_obs)[1]` times.

:::
:::

## Review {#sec-summary_0401}

[↑ top](#)

::: {.callout-note #note-summary_0401}

### Summary

* The POMS data contains detailed information about pollinator species across the UK.
* Here we have also added detailed climate data for each site.
* You have recapped various ways to manipulate data frames in R.

:::

For this part of the practical, the main files you have created are:

- `pollinators.qmd`

These are the files you should select when making your Git commit.


Remember to save and commit your work before moving on to Part 2.

{{< include ../includes/git_commit.qmd >}}


