---
format: html
practical: Practical_4_Statistics
doc-id: "0406"
author: "Katy Brown"
---

```{r setup_0406, include=FALSE}
library(tidyverse)
library(gt)
library(maps)
library(patchwork)
insect_counts <- read_csv("data/insect_counts.csv")
site_metadata <- read_csv("data/site_metadata.csv")
flower_counts <- read_csv("data/flower_counts.csv")
```

# Part 6: Using models to make predictions {#sec-p6_predictions}

::: {.callout-note .partmenu #parts-0406}
## Sections
- @sec-predictions
- @sec-summary_0406
:::

:::{.callout-tip .objectives}
#### Learning objectives
By the end of this part of the practical you will be able to:
* Use a fitted statistical model to predict the expected value of a response variable.
* Explain how a statistical model describes a probability distribution for the response variable.
:::

## Predictions {#sec-predictions}
So far, we have mainly used linear models for hypothesis testing: to ask whether there is evidence that an explanatory variable is associated with a response variable.

However, a fitted model can also be used to predict the expected value of the response variable for particular values of the explanatory variables.

For example, we have already evidence that minimum temperature differs between sites where *H. tumulorum* is present and sites where it is absent.

Therefore, can we use the presence of *H. tumulorum* to predict minimum temperature?

First, let's read our data frame showing presence or absence of *H. tumulorum*.

```{r regenerate_ht_df, message = FALSE}
ht_df <- read_csv("data/ht_df.csv")

gt(head(ht_df))
```
Now we can fit the linear model.

```{r regenerate_ht_lm}
ht_lm <- lm(
  min_temp_coldest_month ~ detected,
  data = ht_df,
)
```

The R function `predict()` can be used to predict the mean value of the response variable under a fitted linear model given a specific value of the explanatory variable.

Let's predict minimum temperature for sites where *H. tumulorum* is present.

```{r predict_temp_ht}
new_expl <- tibble(detected = TRUE)

pred <- predict(
  ht_lm,
  newdata = new_expl,
  interval = "confidence"
)

print(pred)
```
The values in the `fit` column are the predicted minimum temperature with *H. tumulorum* present, and `lwr` (lower) and `upr` (upper) are the lower and upper confidence interval of the prediction.

We can extract the standard deviation from the model (using the function `sigma()`) and plot the predicted distribution. This linear model assumes that observations are normally distributed around the expected value, with a common residual standard deviation

```{r plot_dist_predict_ht}
pred_ht <- rnorm(1000, pred[1, "fit"], sigma(ht_lm))

df_ht <- tibble(data = pred_ht)

ggplot(
  df_ht,
  aes(x = data)
) +
  geom_histogram(bins = 30) +
  theme_bw()
```
The model therefore gives us more than a single predicted value. For each value of the explanatory variable, it estimates the mean of a probability distribution for the response variable. For a linear model, observations are assumed to be normally distributed around this mean, with the residual standard deviation estimated by `sigma()`.

You can make predictions using a linear model with a continuous explanatory variable by using `predict()` in exactly the same way. If there are multiple explanatory variables, the value of each variable needs to be passed to the model, the process is otherwise identical.

::: {.callout-exercise #ex-predict_lm}

In your `pollinators.qmd` Quarto file, fit a linear model to the relationship between annual precipitation (column `annual_total_prec`) and mean annual temperature (column `annual_mean_temp`), with annual precipitation as the response variable, as we did previously in @sec-linear_regression. 

Predict the total annual precipitation and its confidence interval when annual mean temperature is 10°C. Then use the fitted model to simulate and plot the distribution of annual precipitation expected at 10°C.

::: {.callout-answer collapse="true" #ex-predict_lm}

[Fit a linear model to the relationship between annual precipitation (column `annual_total_prec`) and mean annual temperature (column `annual_mean_temp`), with annual precipitation as the response variable]{.blue-text}

```{r predict_lm_ans1}
lm_temp_prec <- lm(annual_total_prec ~ annual_mean_temp, data = site_metadata)

df_unknown <- tibble(annual_mean_temp = 10)

pred_lm <- predict(lm_temp_prec, newdata = df_unknown, interval = "confidence")

print(pred_lm)
```

[Predict the mean and confidence interval and plot the distribution for annual precipitation when annual mean temperature is 10°C.]{.blue-text}

```{r predict_lm_ans2}
pred_prec <- rnorm(1000, pred_lm[1, "fit"], sigma(lm_temp_prec))

df_prec <- tibble(data = pred_prec)

ggplot(
  df_prec,
  aes(x = data)
) +
  geom_histogram(bins = 30) +
  theme_bw()
```
:::
:::
## Review {#sec-summary_0406}

[↑ top](#)

::: {.callout-note #note-summary_0406}

### Summary


:::

For this part of the practical, the main files you have edited are:


These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}

{{< include ../includes/git_push.qmd >}}
