---
title: "temp"
format: html
---
```{r setup_0405, include=FALSE}
library(tidyverse)
library(gt)
library(maps)
library(patchwork)
insect_counts <- read_csv("data/insect_counts.csv")
site_metadata <- read_csv("data/site_metadata.csv")
```

Extension: Can we break a t-test?

A t-test assumes normally distributed populations, but it is often described as robust to violations of this assumption. What does this actually mean?

First, let's simulate two samples of 20 annual mean temperature data points, both with the same mean and standard deviation - let's use the values from our site metadata data frame - a mean of `r mean(pull(site_metadata, annual_mean_temp))` and standard deviation of `r sd(pull(site_metadata, annual_mean_temp))`.

```{r simulate_identical_data}
mean_sites <- mean(pull(site_metadata, annual_mean_temp))
sd_sites <- sd(pull(site_metadata, annual_mean_temp))
n_sites <- 50

sample_1 <- rnorm(n_sites, mean = mean_sites, sd = sd_sites)
sample_2 <- rnorm(n_sites, mean = mean_sites, sd = sd_sites)
```

We can put these two samples into a data frame to more easily make comparisons.

```{r df_identical_data}
sample_df <- tibble(
  sample_1 = sample_1,
  sample_2 = sample_2
) |>
  pivot_longer(
    everything(),
    names_to = "sample",
    values_to = "annual_mean_temp"
  )
```

And summarise the data frame to check the data is as we expected.

```{r check_identical_data}
summary <- sample_df |>
  group_by(sample) |>
  summarise(mean = mean(annual_mean_temp),
            sd = sd(annual_mean_temp))
gt(summary)
```

Both should have approximately the same mean and standard deviation.

Now, let's perform a t-test comparing these two samples - as they have the same mean, the result should not (usually) be significant.

:::{callout-note #note-significant_p}
Note that even when the data perfectly meets the assumptions for a t-test, it is subject to type I error - we expect to see false positives about 5% of the time. If everyone in the course runs this test, the chances are at least a few people will see significant results here.
:::

```{r t_test_identical_data}
t_res <- t.test(annual_mean_temp ~ sample, data = sample_df)
print (t_res$p.value)
```
Let's run a simulation and repeat this process 1,000 times, and see how often we see a p-value < 0.05.

```{r simulate_identical_data2}
t_vec <- c()

for (i in seq_len(1000)) {
  sample_1 <- rnorm(n_sites, mean = mean_sites, sd = sd_sites)
  sample_2 <- rnorm(n_sites, mean = mean_sites, sd = sd_sites)
  sample_df <- tibble(
    sample_1 = sample_1,
    sample_2 = sample_2
  ) |>
    pivot_longer(
      everything(),
      names_to = "sample",
      values_to = "annual_mean_temp"
    )

  t_res <- t.test(annual_mean_temp ~ sample, data = sample_df)$p.value
  t_vec <- c(t_vec, t_res)
}

```


```{r plot_p}
pval_df <- tibble(pvalue = t_vec)

ggplot(
  pval_df,
  aes(x = pvalue)
) +
  geom_histogram(bins = 25)
```
