---
format: html
practical: Practical_4_Statistics
doc-id: "0401"
author: "Katy Brown"
---

```{r setup_0401, include=FALSE}
library(tidyverse)
library(gt)
library(osbng)
library(sf)
library(maps)
insect_metadata <- read_csv("data/insect_metadata.csv")
```

# Part 1: Dataset {#sec-p1_data}

::: {.callout-note .partmenu #parts-0401}
## Sections

- @sec-dataset_pollinators
- @sec-dataset_invertebrates
- @sec-summary_0401
:::

:::{.callout-tip .objectives}
#### Learning objectives
By the end of this part of the practical you will be able to:
:::

We will use three real research datasets throughout both this practical and Practical 6.

## Pollinator dataset {#sec-dataset_pollinators}

The first dataset we will use in this practical is [pan-trap survey data](https://catalogue.ceh.ac.uk/documents/9df83130-e98f-48b0-bbfc-32f5920eedb0) from the UK Pollinator Monitoring Scheme. In this scheme, multiple [pan-traps](https://en.wikipedia.org/wiki/Pan_trap) are set in each of a set of specific 1km squares across the UK. On four days each summer, volunteers will set out traps for six hours, then send the insects to be counted. Flowering plants within a 2m radius of the trap are also recorded.

[comment]: todo_katy link to the preprocessing.

The data from the scheme is provided as data per trap, for this practical we have amalgamated the data across all dates and across each 1km square. The preprocessing of the data is described in XXXX. 

You have been provided with three data files:

* In the **insect counts** CSV file - `data/insect_counts.csv` each row represents a combination of an insect species and a sampling site, and the total number of times that insect was observed at that site. Many of the counts will be zero.

* In the **flower counts** CSV file - `data/flower_counts.csv` each row represents a combination of a flower species and a sampling site, plus TRUE or FALSE depending if that flower was ever observed within 2m of the site.

* In the **site metadata** CSV file - `data/site_metadata.csv`, each row represents a single 1km square sampling site. The data frame contains information about the geographical location, plus land use and climate at the sample site. Additional climate data has been added based on [WorldClim](https://nordata.physics.utoronto.ca/en/annexes-datasets-descriptions/72-precipitation/worldclim-version-2/) data, which has detailed climate data by square averaged from 1970-2000. Various additional statistics have been calculated based on the information in the original data frame.

::: {.callout-note #uk_poms_citation}
**Data citation**

UK Pollinator Monitoring Scheme (2024). 1km square pan trap data from the UK Pollinator Monitoring Scheme, 2017-2021. NERC EDS Environmental Information Data Centre. (Dataset).

The UK Pollinator Monitoring Scheme (PoMS) is a partnership funded jointly by UKCEH and JNCC (through funding from Defra, Scottish Government, Welsh Government, and DAERA). UKCEH’s contribution is funded by the Natural Environment Research Council award number NE/R016429/1 as part of the UK-SCAPE programme delivering National Capability. Between 2017 and 2021, PoMS was funded by UKCEH and Defra (England), Welsh Government, Scottish Government, DAERA (Northern Ireland), and JNCC.
PoMS is indebted to the many volunteers who carry out surveys and contribute data to the scheme.

:::

## Invertebrate dataset

[comment]: todo_katy link to the preprocessing.

In [this publication](https://www.nature.com/articles/s41586-026-10908-7), published in Nature in 2026, the authors investigated the potential of marine invertebrates as a food source, and found these organisms to be very nutrient dense, with the potential to feed billions of people.

They sourced data on nutrient content and various ecological characteristics from numerous organisms, however most of the records are incomplete. In `data/marine_invertebrates.csv` you have been provided with data from the 85 organisms for which data is available for the following measurements:

* `lmax` - maximum length
* `depth_range_deep` - maximum depth
* `FE` - iron mg / 100g
* `CA` - calcium mg / 100g
* `MG` - magnesium mg / 100g
* `order` - taxonomic order

## Septin dataset

Our third dataset is based on [this article](https://doi.org/10.1038/s41467-024-54638-8), published in Nature Communications in 2024. These authors developed a tool to determine the spatial distribution of components inside cells using 3D image analysis. This is based on a measure, $P_n$, of how polar the distribution of a protein is within a cell.

* low $P_n$ → protein distributed relatively symmetrically → less polar
* high $P_n$ → protein concentrated asymmetrically toward one part of the cell → more polar.

Using their tool, they characterised the distribution of 6 members of a family of proteins called Septins, which are involved in cellular compartmentalisation, in haematopoietic stem cells (HSCs) from young mice (aged 10–18 weeks) and aged mice (aged 80+ weeks). They found that most Septins had a lower $P_n$ in aged HSCs, indicating a less polar distribution, and that the organisation of the Septin network was altered with age. This is shown in their Figure 3B.

The data is in your `data/septin.csv` folder.

Choose **at least one** of the following exercises to recap data frame operations.

::: {.callout-exercise #ex_summarise_bees}
In your `pollinators.qmd` Quarto file, using the **pollinator** dataset.

* Read the three CSV files:

  * `data/insect_counts.csv`
  * `data/flower_counts.csv`
  * `data/site_metadata.csv`

* Count how many unique sites there are. Site ID is in the `x1km_square` column of the insect counts data frame and site metadata data frame. 
* Count how many unique observations of insects there are - including multiple observations of the same insect at the same site. Number of observations is in the `n_observed` column of the insect counts data frame.
* Find the insect species which was observed most frequently and how many times it was observed. Insect species is in the column `species` of the insect counts data frame.

::: {.callout-answer #ex_summarise_bees_ans}

[Read the three CSV files]{.blue-text}

```{r summarise_bees_ans1, message=FALSE}
insect_counts <- read_csv("data/insect_counts.csv")
flower_counts <- read_csv("data/flower_counts.csv")
site_metadata <- read_csv("data/site_metadata.csv")
```

[Count how many unique sites there are. Site ID is in the `x1km_square` column of the insect counts data frame and site metadata data frame. ]{.blue-text}

```{r summarise_bees_ans2, message=FALSE}
site_count <- nrow(site_metadata)
print(site_count)
```
There are `r site_count` sites.

[Count how many unique observations of insects there are - including multiple observations of the same insect at the same site. Number of observations is in the `n_observed` column os the insect counts data frame]{.blue-text}

```{r summarise_bees_ans3, message=FALSE}
total_insect_obs <- insect_counts |>
  pull(n_observed) |>
  sum()
print(total_insect_obs)
```
There are `r format(total_insect_obs, scientific=FALSE)` unique insect observations.

[Find the insect species which was observed most frequently and how many times it was observed. Insect species is in the column `species` of the insect counts data frame and the insect metadata data frame.]{.blue-text}

```{r summarise_bees_ans4, message=FALSE}
insect_totals <- insect_counts |>
  group_by(species) |>
  summarise(n_obs = sum(n_observed))

insect_totals <- insect_totals |>
  arrange(desc(n_obs))

gt(head(insect_totals))
```

The most commonly observed species was `r pull(insect_totals, species)[1]`, which was observed `r pull(insect_totals, n_obs)[1]` times.

:::
:::

::: {.callout-exercise #ex_summarise_invertebrates}
In your `invertebrates.qmd` Quarto file, using the **invertebrates** dataset.

* Read the CSV file `data/marine_invertebrates.csv`
* Count how many unique orders of marine invertebrates are present (column `order`)
* Count how many times the CA concentration is greater than the FE concentration.
* Count how many species are in the dataset with the genus *Penaeus*. The function `str_starts(column, string)` will find rows starting with a specific character string.

::: {.callout-answer #ex_summarise_bees_ans}

[Read the CSV file `data/marine_invertebrates.csv`]{.blue-text}

```{r summarise_inverts_ans1, message=FALSE}
inverts <- read_csv("data/marine_invertebrates.csv")
```

[Count how many unique orders of marine invertebrates are present (column `order`)]{.blue-text}
```{r summarise_inverts_ans2, message=FALSE}
n_unique <- inverts |>
  distinct(order) |>
  nrow()
print(n_unique)
```

There are `r n_unique` unique orders.

[Count how many times the CA concentration is greater than the MG concentration.]{.blue-text}

```{r summarise_inverts_ans3, message=FALSE}
mg_ca <- inverts |>
  filter(CA > MG) |>
  nrow()
print (mg_ca)
```
`r mg_ca` rows have a higher CA than MG concentration.

[Count how many species are in the dataset with the genus *Penaeus*.]{.blue-text}

```{r summarise_inverts_ans4, message=FALSE}
pen <- inverts |>
  filter(str_starts(sciname, "Penaeus")) |>
  nrow()
print (pen)
```
`r pen` species are members of the genus *Penaeus*.
:::
:::


## Review {#sec-summary_0401}

[↑ top](#)

::: {.callout-note #note-summary_0401}

### Summary
:::

For this part of the practical, the main files you have created are:


These are the files you should select when making your Git commit.


Remember to save and commit your work before moving on to Part 2.

{{< include ../includes/git_commit.qmd >}}


