---
format: html
practical: Practical_4_Statistics
doc-id: "0406"
author: "Katy Brown"
---

```{r setup_0406, include=FALSE}
library(tidyverse)
library(gt)
library(maps)
library(patchwork)
insect_counts <- read_csv("data/insect_counts.csv")
site_metadata <- read_csv("data/site_metadata.csv")
flower_counts <- read_csv("data/flower_counts.csv")
```

# Part 6: Using models to make predictions {#sec-p6_predictions}

::: {.callout-note .partmenu #parts-0406}
## Sections
- @sec-predictions
- @sec-summary_0406
:::

:::{.callout-tip .objectives}
#### Learning objectives
By the end of this part of the practical you will be able to:

By the end of this part of the practical you will be able to:

* Use a fitted statistical model to predict the expected value of a response variable.
* Interpret a confidence interval around a model prediction.
* Explain how a linear model describes a probability distribution for individual observations around the expected value.
:::

## Predictions {#sec-predictions .core}

[**Core**]{.core-text}

So far, we have mainly used linear models for **hypothesis testing**: to ask whether there is evidence that an explanatory variable is associated with a response variable.

However, a fitted model can also be used to **predict the expected value** of the response variable for particular values of the explanatory variables.

For example, we have already found evidence that minimum temperature differs between sites where *H. tumulorum* is present and sites where it is absent.

Therefore, can we use the presence of *H. tumulorum* to predict minimum temperature?

First, let's read our data frame showing presence or absence of *H. tumulorum*.

```{r regenerate_ht_df, message = FALSE}
ht_df <- read_csv("data/ht_df.csv")

gt(head(ht_df))
```

Now we can fit the linear model.

```{r regenerate_ht_lm}
ht_lm <- lm(
  min_temp_coldest_month ~ detected,
  data = ht_df,
)
```

The R function `predict()` can be used to predict the mean value of the response variable under a fitted linear model given a specific value of the explanatory variable.

Let's predict minimum temperature for sites where *H. tumulorum* is **present**.

```{r predict_temp_ht}
new_expl <- tibble(detected = TRUE)

pred <- predict(
  ht_lm,
  newdata = new_expl,
  interval = "confidence"
)

print(pred)
```

The value in the `fit` column is the predicted mean minimum temperature for sites where *H. tumulorum* is present. `lwr` and `upr` give a confidence interval for this expected mean.

We can extract the standard deviation from the model (using the function `sigma()`) and plot the predicted distribution. This linear model assumes that observations are normally distributed around the expected value, with a common residual standard deviation.

```{r plot_dist_predict_ht}
pred_ht <- rnorm(1000, pred[1, "fit"], sigma(ht_lm))

df_ht <- tibble(data = pred_ht)

ggplot(
  df_ht,
  aes(x = data)
) +
  geom_histogram(bins = 30) +
  theme_bw()
```

The model therefore gives us more than a single predicted value. For each value of the explanatory variable, it estimates the mean of a probability distribution for the response variable. For a linear model, observations are assumed to be normally distributed around this mean, with the residual standard deviation estimated by `sigma()`.

You can make predictions using a linear model with a continuous explanatory variable by using `predict()` in exactly the same way. If there are multiple explanatory variables, the value of each variable needs to be passed to the model, the process is otherwise identical.

::: {.callout-exercise #ex-predict_lm .core}

### Making predictions

[**Core**]{.core-text}

In your `pollinators.qmd` Quarto file, fit a linear model to the relationship between annual precipitation (column `annual_total_prec`) and mean annual temperature (column `annual_mean_temp`), with annual precipitation as the response variable, as we did previously in @sec-linear_regression. 

Predict the mean total annual precipitation and its 95% confidence interval when annual mean temperature is 10°C.

::: {.callout-answer collapse="true" #ex-predict_lm}

[Fit a linear model to the relationship between annual precipitation (column `annual_total_prec`) and mean annual temperature (column `annual_mean_temp`), with annual precipitation as the response variable]{.blue-text}

```{r predict_lm_ans1}
lm_temp_prec <- lm(annual_total_prec ~ annual_mean_temp,
                   data = site_metadata)



```

[Predict the mean total annual precipitation and its 95% confidence interval when annual mean temperature is 10°C.]{.blue-text}

```{r predict_lm_ans2}
df_unknown <- tibble(annual_mean_temp = 10)

pred_lm <- predict(lm_temp_prec,
                   newdata = df_unknown,
                   interval = "confidence")

print(pred_lm)
```
The predicted mean total annual precipitation is `r round(pred_lm[1, "fit"])` mm when annual mean temperature is 10°C. The 95% confidence interval for the expected mean precipitation is [`r round(pred_lm[1, "lwr"])`, `r round(pred_lm[1, "upr"])`] mm.
:::
:::
## Review {#sec-summary_0406}

[↑ top](#)

In this part of the practical, you used fitted linear models to make predictions for new values of explanatory variables. You also considered how a linear model describes not just an expected response, but a distribution of possible observations around that expectation.

::: {.callout-note #note-summary_0406}

### Programming Summary

| Function | What it does |
|---|---|
| `predict()` | Uses a fitted statistical model to predict the expected value of the response variable for specified values of the explanatory variable(s). |
| `sigma()` | Extracts the estimated residual standard deviation from a fitted linear model. |

:::

For this part of the practical, the main files you have edited are:

* `pollinators.qmd`

These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}
