---
title: "Preprocessing Bee Data"
format: html
---

Import relevant libraries

```{r libraries_bee_data, message=FALSE}
library(tidyverse)
library(osbng)
library(sf)
library(terra)
library(rnaturalearth)
library(pollimetry)
```

Read the raw data.

```{r import_bee_data, message=FALSE}
samples <- read_csv("data/ukpoms_1kmpantrapdata_2017-2021_samples.csv")
insects <- read_csv("data/ukpoms_1kmpantrapdata_2017-2021_insects.csv")
flowers <- read_csv("data/ukpoms_1kmpantrapdata_2017-2021_flowers.csv")
```

Check the grid references in the sample table are valid, remove any which are not.

```{r check_grid}
# Convert the text column to BNG reference objects, remove any which are NA
samples <- samples |>
  mutate(grid_ref = as_bng_reference(x1km_square)) |>
  filter(!is.na(grid_ref))
```

Remove extra information in square or round brackets from insect species names and any extra suffix, filter out anything only identified at genus level.

```{r clean_insect_names}
# Remove suffices and anything in brackets from taxon_source
# filter anything without two words remaining
# and put the result in a new column, species
insects <- insects |>
  mutate(
    species = str_extract(
      taxon_source,
      "^[A-Z][a-z]+(?: \\([A-Z][a-z]+\\))? [a-z]+"
    ),
    species = str_remove(
      species,
      " \\([A-Z][a-z]+\\)"
    )
  ) |>
  filter(!is.na(species))

# Same for the flower species
flowers <- flowers |>
  mutate(
    flower_species = str_extract(
      taxon_source,
      "^[A-Z][a-z]+(?: \\([A-Z][a-z]+\\))? [a-z]+"
    ),
    flower_species = str_remove(
      flower_species,
      " \\([A-Z][a-z]+\\)"
    )
  ) |>
  filter(!is.na(flower_species))
```

For each insect species and 1km square, count the number of times each insect occurred, including if it occurred 0 times. 

```{r counts_insects_per_square}
# List the unique insect species
species <- insects |>
  distinct(species)

# List unique combinations of sample ID and site
sample_info <- samples |>
  distinct(sample_id, x1km_square)

# List unique combinations of square and date
survey_events <- samples |>
  distinct(x1km_square)

# Merge the sample info into the insect data frame
insects_plus_samples <- insects |>
  left_join(sample_info, by = "sample_id")

# Count the number of occurrences of each combo of square, species
species_counts <- insects_plus_samples |> count(
  x1km_square,
  species,
  name = "n_observed"
)


# Get every combination of species and sampling events
combos <- survey_events |> crossing(species)

# Merge this with the data frame of the number of occurrences
insects_by_square <- combos |>
  left_join(
    species_counts,
    by = c("x1km_square", "species")
  )

# Replace NA with 0
insects_by_square <- insects_by_square |>
  mutate(n_observed = replace_na(n_observed, 0))

# Add genus column
insects_by_square <- insects_by_square |>
  mutate(genus = word(species, 1))
```

For flowers - for each flower species and 1km square, calculate the total number of floral units observed, including zeros.

```{r flowers_per_square}
# List the unique flower species
flower_species <- flowers |>
  distinct(flower_species)

# Merge the sample info into the insect data frame
flowers_plus_samples <- flowers |>
  left_join(sample_info, by = "sample_id")

# Count the number of occurrences of each combo of square and species
flower_species_counts <- flowers_plus_samples |>
  group_by(x1km_square, flower_species) |>
  summarise(n_observed_flowers = sum(total_floral_units))

# Get every combination of species and sampling events
combos <- survey_events |> crossing(flower_species)

# Merge this with the data frame of the number of occurrences
flowers_by_square <- combos |>
  left_join(
    flower_species_counts,
    by = c("x1km_square", "flower_species")
  )

# Replace NA with 0
flowers_by_square <- flowers_by_square |>
  mutate(n_observed_flowers = replace_na(n_observed_flowers, 0))

flowers_by_square <- flowers_by_square |>
  mutate(flower_genus = word(flower_species, 1))

```

In the samples table, convert Ordance Survey grid reference (BNG reference) locations to latitude and longitude.

```{r convert_refs_bee_data, message=FALSE}
# Convert these to x and y easting and northing references
samples <- samples |> mutate(xy = bng_to_xy(grid_ref, position = "centre"))

# Split x and y
samples <- samples |>
  mutate(
    x = xy[, 1],
    y = xy[, 2]
  )

# Convert these to latitude and longitude on a 4326 map projection
coords <- st_as_sf(
  samples,
  coords = c("x", "y"),
  crs = 27700
) |>
  st_transform(4326)

# Extract the co-ordinates
lonlat <- st_coordinates(coords)

# Put latitude and longitude in separate columns in the samples table
samples <- samples |>
  mutate(
    longitude = lonlat[, 1],
    latitude = lonlat[, 2],
    .before = 3
  )
```

Reduce the number of columns in the sample table to just those we might need. The temperature etc. measurements are not very useful as they're just on that specific day and the volunteers are told to take them on warm days during summer - use WORLDCLIM data instead.

```{r select_bee_data}
samples_sel <- samples |> select(
  country,
  longitude,
  latitude,
  x1km_square,
  land_cover,
  habitat_main,
)
```

Summarise the sample sites per square - where there are multiple samples taken from the same square take the mean of numerical values and the most common of categorical values.

```{r group_samples_bee_data}
sample_summ <- samples_sel |>
  group_by(x1km_square) |>
  summarise(
    country = first(country),
    longitude = first(longitude),
    latitude = first(latitude),
    habitat = names(sort(table(habitat_main), decreasing = TRUE))[1],
    land_cover = names(sort(table(land_cover), decreasing = TRUE))[1]
  )
```

Select the most relevant columns from the insect and flower tables. Add a column to the samples table showing the most common insect family and species at the site.

```{r select_insects_bee_data}
# Select columns and deduplicate
insects_sel <- insects |>
  select(species, english_name, order, family) |>
  distinct(species, english_name, order, family)

# Select columns and deduplicate
flowers_sel <- flowers |>
  select(flower_species, english_name, order, family) |>
  distinct(flower_species, english_name, order, family)

# Find the most common insect family per square
dominant_family <- insects_by_square |>
  inner_join(insects_sel) |>
  group_by(x1km_square, family) |>
  summarise(n = sum(n_observed), .groups = "drop") |>
  group_by(x1km_square) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(x1km_square, dominant_family = family)

insects_sel <- insects_sel |>
  mutate(genus = word(species, 1))

# Find the most common insect species per square
dominant_species <- insects_by_square |>
  group_by(x1km_square, species) |>
  summarise(n = sum(n_observed)) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(x1km_square, dominant_species = species)

insect_summ <- insects_by_square |>
  group_by(x1km_square) |>
  filter(n_observed != 0) |>
  summarize(n_unique = n_distinct(species))

sample_summ <- inner_join(sample_summ, dominant_species) |>
  inner_join(dominant_family) |>
  inner_join(insect_summ)

# Find the most common flower family per square
dominant_flower_order <- flowers_by_square |>
  inner_join(flowers_sel) |>
  group_by(x1km_square, order) |>
  summarise(n = sum(n_observed_flowers), .groups = "drop") |>
  group_by(x1km_square) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(x1km_square, dominant_flower_order = order)

flowers_sel <- flowers_sel |>
  mutate(genus = word(flower_species, 1))

# Find the most common flower species per square
dominant_flower_species <- flowers_by_square |>
  group_by(x1km_square, flower_species) |>
  summarise(n = sum(n_observed_flowers)) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  select(x1km_square, dominant_flower_species = flower_species)

flower_summ <- flowers_by_square |>
  group_by(x1km_square) |>
  filter(n_observed_flowers != 0) |>
  summarize(n_unique_flowers = n_distinct(flower_species))

sample_summ <- inner_join(sample_summ, dominant_flower_species) |>
  inner_join(dominant_flower_order) |>
  inner_join(flower_summ)
```

Add data about insect body size to the insects data frame based on the `pollimetry` R package.

```{r pollimetry}
poll <- tibble(pollimetry_dataset)
poll <- poll |>
  mutate(
    species = Species,
    dry_mass = Spec.wgt,
    intertegular_distance = IT
  ) |>
  select(species, dry_mass, intertegular_distance) |>
  group_by(species) |>
  summarise(
    mean_dry_mass = mean(dry_mass),
    mean_intertegular_distance = mean(intertegular_distance)
  ) |>
  mutate(species = str_replace_all(species, "_", " "))

insects_sel <- insects_sel |> left_join(poll)
```

Summarise the biomass data across sites.

```{r biomass}
insect_biomass <- insects_by_square |> 
  filter(n_observed != 0) |>
  left_join(insects_sel) |>
  filter(!is.na(mean_dry_mass)) |>
  mutate(total_biomass = n_observed * mean_dry_mass) |>
  group_by(x1km_square) |>
  summarise(
    n_insects = sum(n_observed),
    total_insect_biomass = sum(total_biomass),
    mean_insect_mass = total_insect_biomass / n_insects
  )
sample_summ <- sample_summ |> left_join(insect_biomass)
```
Subset habitat types:
```{r subset_habitats}
sample_summ <- sample_summ |>
  mutate(
    broad_habitat = case_when(
      str_detect(habitat, "^H(16|17|18|19|21)") ~ "grassland/herbaceous",
      str_detect(habitat, "^H(25|26|27|32)") ~ "woodland",
      str_detect(habitat, "^H(10|12|13)") ~ "wetland",
      str_detect(habitat, "^H(20|23)") ~ "heathland",
      str_detect(habitat, "^H(35|38)") ~ "arable"
    )
  )


```

Save all the outputs.

```{r save_data}
write_csv(insects_sel, "data/insect_metadata.csv")
write_csv(flowers_sel, "data/flower_metadata.csv")
write_csv(insects_by_square, "data/insect_counts.csv")
write_csv(flowers_by_square, "data/flower_counts.csv")
```

Count the number of unique insect species in each of the selected families per site.

```{r count_dom_fam}
fams <- c("Apidae", "Halictidae", "Syrphidae")

by_spp_met <- inner_join(species_counts, insects_sel) |>
  filter(family %in% fams)

counts <- by_spp_met |>
  group_by(x1km_square, family) |>
  count() |>
  ungroup() |>
  complete(
    x1km_square,
    family = fams,
    fill = list(n = 0)
  )  |>
  pivot_wider(
    names_from = family,
    values_from = n,
    names_prefix = "n_",
    values_fill = 0
  ) |>
  rename_with(tolower)

sample_summ <- inner_join(sample_summ, counts)
```

Read the [WorldClim](https://nordata.physics.utoronto.ca/en/annexes-datasets-descriptions/72-precipitation/worldclim-version-2/) data, which has climate data by square averaged from 1970-2000. These are 10 arc-minutes squares but should be close enough for our purposes.

Take the [19 bioclim variables](https://www.nccs.nasa.gov/bioclim/) plus temperature for each month and elevation.

Data was originally downloaded using the package `worldclim`: https://search.r-project.org/CRAN/refmans/geodata/html/worldclim.html on 17/08/26.

```{r read_worldclim}
# Find worldclim bioclim data
bio_files <- list.files(
  "/home/katy/WorldClim_data_unzipped/",
  pattern = ".*bio.*.tif$",
  full.names = TRUE
)

# Find worldclim temperature data
temp_files <- list.files(
  "/home/katy/WorldClim_data_unzipped/",
  pattern = ".*tavg.*.tif$",
  full.names = TRUE
)

# Find worldclim elevation data
elev_files <- list.files(
  "/home/katy/WorldClim_data_unzipped/",
  pattern = ".*elev.*.tif$",
  full.names = TRUE
)

# Get the longitude and latitude of the sample sites
sites <- distinct(select(sample_summ, c(x1km_square, longitude, latitude)))

# Convert to a vector, match the projection to the WC data.
sites_vect <- vect(
  sites,
  geom = c("longitude", "latitude"),
  crs = "EPSG:4326"
)

bioclim <- rast(bio_files)
bio_values <- extract(bioclim, sites_vect)

bioclim_temp <- rast(temp_files)
temp_values <- extract(bioclim_temp, sites_vect)

bioclim_elev <- rast(elev_files)
elev_values <- extract(bioclim_elev, sites_vect)
```

Clean this data - pick the variables we are interested in, rename so the names are informative.

```{r tibble_worldclim}
# Convert to data frames
world_clim <- tibble(bio_values)
world_temp <- tibble(temp_values)
world_elev <- tibble(elev_values)

# Create two variables directly - number of
# months >= 15C and elevation.
months_over_15 <- rowSums(world_temp >= 15)
elevation <- pull(world_elev, wc2.1_10m_elev)

world_clim <- world_clim |>
  mutate(
    x1km_square = sites_vect$x1km_square,
    annual_mean_temp = wc2.1_10m_bio_1,
    max_temp_warmest_month = wc2.1_10m_bio_5,
    min_temp_coldest_month = wc2.1_10m_bio_6,
    temp_annual_range = wc2.1_10m_bio_7,
    annual_total_prec = wc2.1_10m_bio_12,
    total_prec_wettest_month = wc2.1_10m_bio_13,
    total_prec_driest_month = wc2.1_10m_bio_14
  ) |>
  select(
    x1km_square,
    annual_mean_temp,
    max_temp_warmest_month,
    min_temp_coldest_month,
    temp_annual_range,
    annual_total_prec,
    total_prec_wettest_month,
    total_prec_driest_month
  ) |>
  mutate(
    n_months_at_least_15C = months_over_15,
    elevation = elevation
  )
```

Calculate the distance to the coast from each point.

```{r coast_distance}
# Get the position of the UK coastline - from the rnaturalearth package
coast <- ne_coastline(
  scale = "medium",
  returnclass = "sf"
) |>
  st_crop(
    xmin = -10,
    xmax = 4,
    ymin = 49,
    ymax = 61
  ) |>
  st_transform(27700)

# Transform the site co-ordinates into the same co-ordinate system
# as the coastline
sites_sf <- sample_summ |>
  st_as_sf(
    coords = c("longitude", "latitude"),
    crs = 4326
  ) |> st_transform(27700)

# Calculate the distance between each point and each section of
# coastline

distance_to_coast <- st_distance(
  sites_sf,
  coast
)

# Find the shortest, convert to km
distance_to_coast_km <- apply(
  distance_to_coast,
  1,
  min
) |>
  as.numeric() / 1000
```

Merge the site data with the worldclim data and save.

```{r export_samples}
sample_summ <- sample_summ |> full_join(world_clim) |>
  mutate(distance_to_coast = distance_to_coast_km)
write_csv(sample_summ, "data/site_metadata.csv")
```
