---
title: "Filter Ensembl Data"
author: "Katy Brown"
---
Import the biomaRt library

```{r library_biomart}
library(biomaRt)
library(tidyverse)
library(gt)
```

Set up the "mart" - the R object used to extract the data from the Biomart database.

```{r use_mart}
mart <- useMart(
  "ensembl",
  dataset = "hsapiens_gene_ensembl"
)
```

For all genes with the biotype `protein_coding`, retrieve the hgnc_symbol, ensembl_gene_id, ensembl_transcript_id and whether the transcript is canonical i.e. Ensembl's chosen representative transcript for the gene.

```{r get_data}
all_transcripts <- getBM(
  attributes = c(
    "hgnc_symbol",
    "ensembl_gene_id",
    "ensembl_transcript_id",
    "transcript_is_canonical"
  ),
  filters = "biotype",
  values = "protein_coding",
  mart = mart
)
```

Filter to get the canonical transcripts with a HGNC ID only.

```{r filter_canonical}
# Convert to a tibble
all_transcripts <- tibble(all_transcripts)

# Filter
canonical_transcripts <- all_transcripts |>
  filter(transcript_is_canonical == 1,
         hgnc_symbol != "")
```

Extract the transcript IDs into a vector.

```{r get_transcript_ids}
transcript_ids <- canonical_transcripts |>
  pull(ensembl_transcript_id)
```

Pull the 'cds_length' for these IDs from Biomart.

```{r cds_length}
cds_lengths <- getBM(
  attributes = c(
    "ensembl_transcript_id",
    "cds_length"
  ),
  filters = "ensembl_transcript_id",
  values = transcript_ids,
  mart = mart
)
```

Join these with the gene names.

```{r join_dfs}
cds_lengths <- tibble(cds_lengths)
joined <- inner_join(canonical_transcripts, cds_lengths)
```

Keep only the relevant columns and deduplicate (on the hgnc_symbol column only because for some reason a few have two canonical transcripts).

(dplyr::select is to specify to use the dplyr select function and not the biomart one)

```{r select_cols}
genes_cds_lengths <- joined |>
  dplyr::select(hgnc_symbol, cds_length) |>
  distinct(hgnc_symbol, .keep_all = TRUE)
```

Rename the columns to match our other data.

```{r rename}
genes_cds_lengths <- genes_cds_lengths |> rename(Gene_ID = hgnc_symbol)
```

Check it looks OK.

```{r view}
gt(head(genes_cds_lengths))
```

Save the data frame.

```{r save df}
write_csv(genes_cds_lengths, "data/gene_cds_lengths.csv")
```
