---
format: html
practical: Practical_1_Introduction_to_R
doc-id: "0104"
author: "Katy Brown"
---

```{r setup0104, include=FALSE}
library(gt)
```

# Part 4: Vectors, data frames and plots {#sec-p4_vectors_data_frames_plots}

::: {.callout-note .partmenu #parts-0104}
## Sections
- @sec-packages
- @sec-vectors
- @sec-data_frames
- @sec-csv
- @sec-plots
- @sec-summary_0104
:::

:::{.callout-tip .objectives #objectives-0104}
## Learning objectives

By the end of this part of the practical you should be able to:

- Install and load a package in R.
- Create and use vectors in R.
- Access individual elements of a vector using an index.
- Create and view a simple data frame.
- Recognise and read data stored in CSV format.
- Create a simple line plot in R using ggplot2.

:::

## Packages {#sec-packages .core}

[↑ top](#)

[**Core**]{.core-text}

R comes with many useful functions, but additional functions are available in packages. Packages are collections of functions and other resources designed for particular purposes.

Normally, you need to install a package **once** before you can use it, and load it **each time you start a new R session** in which you want to use it.

Let's install the package `fortunes`, to demonstrate.

```{r install_fortunes}
install.packages("fortunes")
```
We can then load the package into R using the function `library()`:

```{r load_fortunes}
library(fortunes)
```

This package just includes one function, `fortune()`, which outputs a random quote about R.

```{r use_fortunes}
fortune()
```

You have already installed all the packages you will need for this practical by following the instructions in @sec-install.

However, we will sometimes need to load a package, for example, for this practical we will use some functions from the [`tidyverse`](https://tidyverse.org/) package collection for data science and from the [`gt`](https://gt.rstudio.com/) package for visualising data frames.

Therefore, you need to include the following code at the beginning of your Quarto document.

::: {.callout-note .copycode #copycode-load_tidyverse}

### Include this code

```{r load_tidyverse, message=FALSE}
library(tidyverse)
library(gt)
```

:::

::: {.callout-note #note-copy_code}

### Copying code

From here onwards, if you need to include code in your Quarto document for future exercises to work, it will be highlighted with "Include this code" as above.
:::

## Vectors  {#sec-vectors .core}

[↑ top](#)

All of the variables we've created so far just have a single element - one number, one section of text, or one TRUE/FALSE.

However, we can also create variables which contain multiple elements.

One of the most common ways of storing multiple values in R is a vector.

Vectors in R are (usually) specified using the function `c()`.

```{r first_vector}
my_first_vector <- c(1, 5, 9)
print(my_first_vector)
```

You can have vectors with all numbers, all text, or all logical values, but everything needs to be the same type.

```{r text_bool_vectors}
character_vec <- c("hello", "this", "is", "a", "character", "vector")
print(character_vec)

logical_vec <- c(TRUE, FALSE, TRUE)
print(logical_vec)
```

If you have existing variables, you can put them into a vector using their variable names.

```{r vectors_of_variables}
x <- 1
y <- 2
z <- 3

xyz_vec <- c(x, y, z)
```

::: {.callout-exercise #ex-vectors1 .core}

Let's look again at our dataset in which you have treated one batch of cells with radiation and kept the other as a control, then counted the number of live cells before and after.

The attributes of your two batches of cells are in this table:

```{r table_cells, echo=FALSE}
df <- data.frame(
  cell_ID = c("CX1", "CX2"),
  treated = c(TRUE, FALSE),
  num_cells_before = c(5415, 8123),
  num_cells_after = c(1432, 7543)
)

gt(df)
```

In your existing `prac1_exercises.qmd` Quarto document, combine the variables you made earlier into four vectors, one for each column of the table.

::: {.callout-answer collapse="true" #ex-vectors1_ans}
```{r vectors1_ans}
# You should already have this part
cell_id_1 <- "CX1"
cell_id_2 <- "CX2"

treated_1 <- TRUE
treated_2 <- FALSE

n_before_1 <- 5415
n_before_2 <- 8123

n_after_1 <- 1432
n_after_2 <- 7543

# Combine these into vectors
cell_id <- c(cell_id_1, cell_id_2)
treated <- c(treated_1, treated_2)
n_before <- c(n_before_1, n_before_2)
n_after <- c(n_after_1, n_after_2)
```
:::
:::

### Accessing elements of a vector

Once our values are in a vector, we can access them using an *index*, which refers to the position of the value within the vector. Indices are indicated within square brackets - `[]`.

::: {.callout-note #note-one_based}

### One-based indexing

If you've previously used a language such as Python, you'll notice that indexing in R is 1-based - so the first element in a vector is element 1 - rather than 0-based.

:::

```{r vector_indices}
nice_food <- c("cheese", "beans", "potatoes")

print(nice_food[1])

print(nice_food[2])

print(nice_food[3])
```


::: {.callout-exercise #ex-index .core}

### Indexing

[**Core**]{.core-text}

In your `prac1_exercises.qmd` Quarto file, isolate the fourth value in this vector and assign it to a new variable, then print its value.

```{r vector_indices2}
all_my_pets <- c("dog", "horse", "guinea pig", "chicken", "cat")
```

::: {.callout-answer collapse="true" #ex-index_ans}

```{r index_ans}
# for example
all_my_pets <- c("dog", "horse", "guinea pig", "chicken", "cat")
my_only_bird <- all_my_pets[4]
print(my_only_bird)
```
:::
:::

### Vectorisation {#sec-vectorisation .fold .extension}

[**Extension**]{.extension-text}

One useful feature of R is **vectorisation**. Many R operations can be applied to an entire vector at once.

For example, adding 10 to a vector adds 10 to every element:

```{r vectorisation}
x <- c(1, 2, 3, 4, 5)
print(x + 10)
```
Similarly, functions will often operate on every element of a vector:

```{r vectorisation_functions}
print(sqrt(x))
```
::: {.callout-exercise #ex-vectorisation .extension}

Create a vector as follows:

```{r vec_ex_1}
x <- c(2, -4, 6, -2)
```

Before running the code, predict the output of each of the following expressions. Then evaluate them in R.

```{r vec_ex_2}
x * 2
x > 5
sum(x)
abs(x)
```
:::

## Data Frames {#sec-data_frames .core}

[↑ top](#)

[**Core**]{.core-text}

Our next variable type is a **data frame**, which is one of the types of data structure you'll most commonly use. A data frame is basically a table, consisting of rows and columns. Each row is a single record, corresponding to some entity - a cell, a person, a country - and each column is a specific piece of information.

For example:

```{r csv_example, echo=FALSE}
df <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(df)
```

In this data frame, there are three columns - fruit, length and weight, and three rows, each containing information about a specific fruit.

We can build a data frame using the R function `tibble()` (note that there is also an R function `data.frame()` which you might see referenced elsewhere, usually modern applications use `tibble()` as it is more user-friendly).

I would specify the data frame above as follows:

```{r make_tibble}
my_data_frame <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)


print(my_data_frame)
```

Although we can see our data frame using the `print()` function, it's not formatted particularly nicely. Instead, we can use the function `gt()` to display a nicely formatted version of our data frame in our Quarto document.

```{r gt}
my_data_frame <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(my_data_frame)
```

::: {.callout-exercise #ex-celldf .core}

### Creating a data frame

[**Core**]{.core-text}

Let's revisit our radiation experiment data.

```{r table_cells_df, echo=FALSE}
df <- data.frame(
  cell_ID = c("CX1", "CX2"),
  treated = c(TRUE, FALSE),
  num_cells_before = c(5415, 8123),
  num_cells_after = c(1432, 7543)
)

gt(df)
```

In your `prac1_exercises.qmd` Quarto file, using a data frame and the `tibble` function, reproduce this table, then visualise it using the `gt` function.

::: {.callout-answer collapse="true" #ex-celldf_ans}

```{r celldf_ans}
# for example
df <- tibble(
  cell_ID = c("CX1", "CX2"),
  treated = c(TRUE, FALSE),
  num_cells_before = c(5415, 8123),
  num_cells_after = c(1432, 7543)
)

gt(df)
```
:::
:::

## CSV Format  {#sec-csv .core}

[↑ top](#)

[**Core**]{.core-text}

The data we will use in many of our practicals is stored in comma-separated values (CSV) format, a plain-text file format where each line represents a row and values in columns are separated by commas.

For example, for this table:

```{r csv_example2, echo=FALSE}
df <- data.frame(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(df)
```


The contents of the CSV file would be:
```text
fruit,length,weight
apple,22,150
pear,30,165
banana,55,170
```

There is a file provided in your `data` folder for Practical 1 named `weather_data.csv`. Open this file on your computer in a text editor (for example NotePad, TextEdit) and examine it.

::: {.callout-exercise #ex-weather .core}

### Examining a CSV file

[**Core**]{.core-text}

Based on this table, what was the temperature in degrees C on 2026-07-02 at 12:00:00?

::: {.callout-answer collapse="true" #ex-weather_ans}
17.8
:::

:::


### Reading CSV Files {#sec-reading_csv .core}

The R function `read_csv` reads a CSV file into R. 

If I wanted to read the file `weather_data.csv`, the code I would use would be as follows. I can view the data frame with the same `gt` function we used previously.

```{r read_weather_data}
weather <- read_csv("data/weather_data.csv")
gt(weather)
```

::: {.callout-exercise #ex-public_health .core}

### Reading UK Public Health Monitoring Data

[**Core**]{.core-text}

The UK government provides some public health data for England (other UK countries have data available elsewhere) on their data portal, here - https://ukhsa-dashboard.data.gov.uk/. The data is available for download in CSV format.

Have a look through the datasets with measurements taken over time which are available on this page. Choose a category, then select a dataset that is displayed as a line plot with a **single line**. Download the CSV formatted data to your computer. You can find these by clicking on the graph then choosing "Download", then "CSV format". Move this file into the `{{< meta practical >}}/data` folder of your `{{< meta git_repo >}}`  folder

::: {.callout-note}
If you have difficulty downloading a dataset, use the example rhinovirus dataset provided in the data folder as `data/ukhsa-chart-download-rhinovirus.csv`.
:::

In a new Quarto report, `{{< meta practical >}}/ukhsa_data_analysis.qmd`, load this table into R and then view the resulting table with `gt()`. These tables are mostly long, so we can use the `head()` function together with `gt()` to just visualise the first ten lines:

For example:

```{r gt_head}
rhinovirus <- read_csv("data/ukhsa-chart-download-rhinovirus.csv")

gt(head(rhinovirus))
```

Don't forget to write a line or two describing what you did - for example, which dataset you have selected.

You will need this section at the top of your Quarto document:

::: {.callout-note .copycode #copy-load_tv_gt}

```{r load_tidyverse2}
library(tidyverse)
library(gt)
```

:::

### Downloading data via an API {#sec-api .bonus}

Often, when we want to work with data which is regularly updated, we will use an application programming interface (API) to directly access data programmatically. UKHSA provides an API for accessing the data underlying its dashboard. 

The UKHSA API is organised hierarchically. Navigating its hierarchical structure takes a little getting used to.

To find a dataset, we need to work through several levels:

`theme` → `sub_theme` → `topic` → `geography_type` → `geography` → `metric`

Start by opening this URL in your browser:

https://api.ukhsa-dashboard.data.gov.uk/themes/

Here, you need to navigate through the **theme**, then **sub-theme**, then **topic**, **geography_type**, **geography**. This should lead you to a list of metrics, each with a `name` and a `link`. The names of these metrics are not exactly intuitive, but you can find out what they mean by searching for them by `name` [here](https://ukhsa-dashboard.data.gov.uk/metrics-documentation). 

For example, I have chosen *Clostridioides. difficile* monthly cases by onset type, for which the link is:

`https://api.ukhsa-dashboard.data.gov.uk/themes/infectious_disease/sub_themes/gastrointestinal/topics/C-difficile/geography_types/Nation/geographies/England/metrics/c-difficile_cases_countsByOnsetType`

Now we can access the data via the API. 

We need to import a package, `httr2`.

```{r api_hsa_packages}
library("httr2")
```

We can also assign our URL to a variable.

```{r api_url_variable}
url <- "https://api.ukhsa-dashboard.data.gov.uk/themes/infectious_disease/sub_themes/gastrointestinal/topics/C-difficile/geography_types/Nation/geographies/England/metrics/c-difficile_cases_countsByOnsetType"
```

We need to append this to our URL - `?page_size=365`. You have probably seen similar additions to URLs before, which tell a website more specifically what you want it to show you. These are called query parameters. Here, page_size=365 tells the UKHSA API to return up to 365 records.

::: {.callout-note #note-api_page_limit}

The API allows a maximum of 365 records per page. Some datasets contain more than 365 records, so this may not retrieve the complete dataset. For this exercise, it does not matter.

:::

You can concatenate strings in R using the function `paste0`, as follows:

```{r api_url_variable2}
url <- paste0(url, "?page_size=365")
print(url)
```

Now we are ready to import our data. First, we structure a request for the data from the URL, using the function `request()`.

```{r api_request}
api_request <- request(url)
```

Next, we use the function `req_perform()` to perform the request we've built.

```{r api_perform}
response <- req_perform(api_request)
```

The returned data is encoded in [JSON format](https://en.wikipedia.org/wiki/JSON), a common and useful data format used for storing and transferring data. The function `resp_body_json()` reads this JSON data into an R object.

```{r api_json}
json_response <- resp_body_json(response)
```

All of our data is now in this object. We can describe the object using the R function `str()`, which displays the structure of an R object. The argument `max.level = 1` tells R to only show the top level of information (otherwise it will show us something for every data frame row).

```{r api_json_str}
str(json_response, max.level = 1)
```

The part we are interested in is contained in `$results` - this is a list of observations, which will become the rows in our data frame. We haven't encountered R lists yet. Lists are a flexible way of storing multiple objects in R, but they are trickier than vectors, so we won't cover them until Practical 6.

We don't want our results to be in a list, we want a data frame. We can convert a list to a data frame using the function `bind_rows()`.

```{r api_json_make_df}
results_df <- bind_rows(json_response$results)
gt(head(results_df))
```

We should now have our data frame, formatted exactly as it would be if you'd downloaded it directly.

There is one additional step - for our plot later it will be easier if there is a single data series to plot. Some UKHSA metrics contain multiple data series, for example for different strata, sexes or age groups. Before plotting our data, we can check whether any of these columns contain more than one value:

We can check by running the function `unique()` on the data in the `stratum`, `sex` and `age` columns. The function `pull()` extracts a single column from the data frame.

```{r unique_api_stratum}
print(unique(pull(results_df, stratum)))
```

```{r unique_api_sex}
print(unique(pull(results_df, sex)))
```
```{r unique_api_age}
print(unique(pull(results_df, age)))
```

I have several strata but only one age and sex. I can filter to include only one stratum by using the `filter` function on my data frame.

```{r filter_api}
filt <- filter(results_df, stratum == "Total cases")
gt(filt)
```

This may seem like a lot of work to do something you could do manually in two minutes! The advantage of an API is that the process can now be automated. The same code can retrieve newly updated data without somebody having to visit the website and download a new file each time. Programmatic data retrieval can also make an analysis more reproducible and reduce opportunities for manual error. For example, during COVID many websites pulled data from the UKHSA website to display or otherwise use specific aspects of the data.

:::{.callout-exercise #ex-api .bonus}

In your `ukhsa_data_analysis.qmd` Quarto file, download a metric of your choice and convert it to a data frame. Your graph later will be more interesting if you choose something measured by week or by day, rather than by year, as most records start less than 10 years ago.

Check the `stratum`, `sex` and `age` columns. If any contain multiple values, use `filter()` to select a single data series.

:::{.callout-answer collapse="true" #ex-api_ans}
```{r api_ans}
# Assign URL
url <- "https://api.ukhsa-dashboard.data.gov.uk/themes/infectious_disease/sub_themes/gastrointestinal/topics/C-difficile/geography_types/Nation/geographies/England/metrics/c-difficile_cases_countsByOnsetType"

# Add page size
url <- paste0(url, "?page_size=365")

# Set up the request
api_request <- request(url)

# Perform the request
response <- req_perform(api_request)

# Convert to JSON
json_response <- resp_body_json(response)

# Convert to a data frame
results_df <- bind_rows(json_response$results)

# Filter to a single value of stratum
filt <- filter(results_df, stratum == "Total cases")

# Display
gt(filt)
```
:::
:::

## Plotting in R {#sec-plots .core}

[↑ top](#)

[**Core**]{.core-text}

### Initiating a Plot {#sec-initiating_plots .core}

We will now learn to draw some basic graphs in R. We will learn much more about this over future weeks.

We start a new, empty plot using the function `ggplot`. The input for this function is our data frame variable, which contains the data we are going to plot.

In this example, I'll plot data from the `weather` data frame.

```{r load_weather, echo=FALSE}
gt(weather)
```

First, I read the table and initiate a blank canvas on which to put my plot.

```{r initiate_plot}
weather <- read_csv("data/weather_data.csv")

ggplot(weather)
```


### Adding axes {#sec-adding_axes .core}

We use another function, `aes()`, as an argument to the `ggplot()` function,  telling R how various attributes of the data map to the "aesthetics" of the plot.

For example, here we choose which data we will represent with the *x* and *y* axes.

```{r axes_plot}
ggplot(weather, aes(
  x = observation_time,
  y = temperature_c
))
```

### Adding Data {#sec-adding_data .core}

Now we can add some data to our plot. In `ggplot2`, we build plots by adding layers using the + symbol.

Here, we add a line representing the data using the `geom_line()` function.

```{r line_plot}
ggplot(weather, aes(
  x = observation_time,
  y = temperature_c
)) +
  geom_line()
```

### Adding extra features

We can add some additional features here to make the plot look nicer. For example, here I will change the colour of the line to red, add *x* and *y* axis labels and add a title.

```{r colour_line_plot}
ggplot(weather, aes(
  x = observation_time,
  y = temperature_c
)) +
  geom_line(color = "red") +
  xlab("Date") +
  ylab("Temp (C)") +
  ggtitle("Temperature by Date")
```

::: {.callout-exercise #ex-public_health_plot .core}

### Plotting public health data

[**Core**]{.core-text}

Following on from the UKHSA public health data you downloaded above and loaded into R, in the same Quarto document:

* Plot a line graph, where the x axis is date (contained in the `date` column), and the *y* axis is the metric measured in the table (from the `metric_value`) column.

::: {.callout-answer collapse="true" #ex_public_health_plot_ans}

For example:

```{r public_health_plot_ans1}
ggplot(rhinovirus, aes(
  x = date,
  y = metric_value
)) +
  geom_line()
```
:::

::: {.callout-exercise #ex-public_health_plot2 .rec}

### Styling your plot

[**Recommended**]{.rec-text}

Edit the code you wrote above to:

* Change the colour of the line and add appropriate axis labels and a title - you can go back to the UKHSA data website to see exactly what is being measured.

::: {.callout-answer collapse="true" #ex_public_health_plot_ans2}

For example:

```{r public_health_plot_ans2}
ggplot(rhinovirus, aes(
  x = date,
  y = metric_value
)) +
  geom_line(color = "red") +
  xlab("Date") +
  ylab("Percentage") +
  ggtitle("Percentage of positive PCR test results for rhinovirus")
```
:::
:::

::: {.callout-exercise #ex-pretty_plot .extension}

### Make your plot look good

[**Extension**]{.extension-text}

The plot you have created contains the information we need, but there are many ways we could improve its appearance.

Explore the [ggplot2 documentation](https://ggplot2.tidyverse.org/) and improve your plot. For example, you could:

- Try a different **theme**.
- Change the appearance of the line.
- Change the size and style of text.
- Adjust the appearance of the axes or grid lines.

Choose changes that you think make your plot clearer or more visually appealing. You don't need to use all of the suggestions above.

:::


## Review {#sec-summary_0104}

[↑ top](#)

::: {.callout-note #note-summary_0104}
## Summary

* Additional functions are available in **packages**, which must be installed and loaded before they can be used.
* Multiple values of the same type can be combined into a **vector** using the function `c()`.
* Individual elements of a vector can be accessed using their **index** in square brackets (`[]`).
* Tabular data can be represented using a **data frame**; we can create data frames using `tibble()`.
* Data is commonly made available in **comma-separated values (CSV)** format.
* `read_csv()` reads data from a CSV file into an R data frame.
* We can create plots using **ggplot2**, mapping variables to features of the plot using `aes()`.
:::

For this part of the practical, the main files you have created are:

- `prac1_exercises.qmd`
- `ukhsa_data_analysis.qmd`
- the UKHSA CSV file you downloaded into the `data` folder

These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}

{{< include ../includes/git_push.qmd >}}
