---
format: html
---

```{r setup, include=FALSE}
library(gt)
```

# Part 4: Vectors, data frames and plots {#sec-p4_vectors_dataframes_plots}

::: {.callout-note .partmenu}
## Sections
- @sec-packages
- @sec-vectors
- @sec-dataframes
- @sec-csv
- @sec-plots
- @sec-summary_0104
:::

:::{.callout-tip .objectives}
## Learning objectives

By the end of this part of the practical you should be able to:

* Install and load a package in R
* Recognise and use the vector variable type.
* Isolate elements of the vector using an index.
* Create and view a simple data frame
* Recognise and use the CSV file type
* Plot a simple scatter plot in R

:::

## Packages {#sec-packages}

[↑ top](#)

The standard version of R contains many useful packages and functions for data analysis. However, there are also thousands of additional **packages** containing additional functions for specific purposes.

Normally, you need to install these packages in RStudio before you use them.

Let's install the package `fortunes`, to demonstrate.

```{r install_fortunes}
install.packages("fortunes")
```
We can then load the package into R using the function `library`:

```{r load_fortunes}
library(fortunes)
```

This package just includes one function, `fortune`, which outputs a random quote about R.

```{r use_fortunes}
fortune()
```
You have already installed all the packages you will need for this practical by following the instructions in @sec-install.

However, we will sometimes need to load a package, for example, for this practical we will use some functions from the [`tidyverse`](https://tidyverse.org/) package collection for data science and from the [`gt`](https://gt.rstudio.com/) package for visualising data frames.

Therefore, you need to include the following code at the beginning of your Quarto document.

```{r load_tidyverse}
library(tidyverse)
library(gt)
```

## Vectors  {#sec-vectors}

[↑ top](#)

All of the variables we've created so far just have a single element - one number, one section of text, or one TRUE/FALSE.

However, we can also create variables which contain multiple elements.

One of the most common types of multi-element variable in R is a **vector**.

Vectors in R are (usually) specified using the function `c`.

```{r first_vector}
my_first_vector <- c(1, 5, 9)
print(my_first_vector)
```

You can have vectors with all numbers, all text, or all logical values, but everything needs to be the same type.

```{r text_bool_vectors}
character_vec <- c("hello", "this", "is", "a", "character", "vector")
print(character_vec)

logical_vec <- c(TRUE, FALSE, TRUE)
print(logical_vec)
```

If you have existing variables, you can put them into a vector using their variable names.

```{r vectors_of_variables}
x <- 1
y <- 2
z <- 3

xyz_vec <- c(x, y, z)
```

::: {.callout-exercise #ex_vectors1}

{{< fa solid anchor >}}

Let's look again at our dataset in which you have treated one batch of cells with radiation and kept the other as a control, then counted the number of live cells before and after.

The attributes of your two batches of cells are in this table:

```{r table_cells, echo=FALSE}
df <- data.frame(
  cell_ID = c("CX1", "CX2"),
  treated = c(TRUE, FALSE),
  num_cells_before = c(5415, 8123),
  num_cells_after = c(1432, 7543)
)

gt(df)
```

Combine the variables you made earlier into four vectors, one for each column of the table.

::: {.callout-answer collapse="true"}
```{r vector_df_solution}
# for example
cell_id_1 <- "CX1"
cell_id_2 <- "CX2"

treated_1 <- TRUE
treated_2 <- FALSE

n_before_1 <- 5415
n_before_2 <- 8123

n_after_1 <- 1432
n_after_2 <- 7543

cell_id <- c(cell_id_1, cell_id_2)
treated <- c(treated_1, treated_2)
n_before <- c(n_before_1, n_before_2)
n_after <- c(n_after_1, n_after_2)
```
:::
:::

### Accessing elements of a vector

Once our values are in a vector, we can access them using an *index*, which refers to the position of the value within the vector. Indices are indicated within square brackets - `[]`.

```{r vector_indices}
nice_food <- c("cheese", "beans", "potatoes")

print(nice_food[1])

print(nice_food[2])

print(nice_food[3])
```


### Index Exercise 

::: {.callout-exercise #ex_vectors1}

{{< fa solid anchor >}}

Isolate the fourth value in this vector and assign it to a new variable, then print its value.

```{r vector_indices_exercise, exercise=TRUE}
all_my_pets <- c("dog", "horse", "guinea pig", "chicken", "cat")
```

::: {.callout-answer collapse="true"}
```{r vector_indices_exercise_solution, exercise=TRUE}
# for example
all_my_pets <- c("dog", "horse", "guinea pig", "chicken", "cat")
my_only_bird <- all_my_pets[4]
print(my_only_bird)
```
:::
:::

## Data Frames {#sec-dataframes}

[↑ top](#)

Our next variable type is a **data frame** (sometimes called a **tibble**), which is one of the types of variable you'll most commonly use. A data frame is basically a table, consisting of rows and columns. Each row is a single record, corresponding to some entity - a cell, a person, a country - and each column is a specific piece of information.

For example:

```{r csv_example, echo=FALSE}
df <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(df)
```

In this data frame, there are three columns - fruit, length and weight, and three rows, each containing information about a specific fruit.

We can build a data frame using the R function `tibble` (note that there is also an R function `data.frame` which you might see referenced elsewhere, usually modern applications use `tibble` as it is more user-friendly).

I would specify the data frame above as follows:

```{r make_tibble}
my_dataframe <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)


print(my_dataframe)
```

Although we can see our data frame using the `print()` function, it's not formatted particularly nicely. Instead, we can use the function `gt`.

```{r gt}
my_dataframe <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(my_dataframe)
```

::: {.callout-exercise #ex_celldf}

{{< fa solid flask >}}

Let's revisit our radiation experiment data.


```{r table_cells_df, echo=FALSE}
df <- data.frame(
  cell_ID = c("CX1", "CX2"),
  treated = c(TRUE, FALSE),
  num_cells_before = c(5415, 8123),
  num_cells_after = c(1432, 7543)
)

gt(df)
```

Using a data frame and the `tibble` function, reproduce this table, then visualise it using the `gt` function.

::: {.callout-answer collapse="true"}

```{r df_exercise_solution}
# for example
df <- tibble(
  cell_ID = c("CX1", "CX2"),
  treated = c(TRUE, FALSE),
  num_cells_before = c(5415, 8123),
  num_cells_after = c(1432, 7543)
)

gt(df)
```
:::
:::

## CSV Format  {#sec-csv}

[↑ top](#)

The data we will use in many of our practicals is stored in comma-separated values (CSV) format, a plain-text file format where each line represents a row and values in columns are separated by commas.

For example, for this table:

```{r csv_example2, echo=FALSE}
df <- data.frame(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(df)
```


The contents of the CSV file would be:
```text
fruit,length,weight
apple,22,150
pear,30,165
banana,55,170
```

There is a file provided in your `data` folder for Practical 1 named `weather.csv`. Open this file on your computer in a text editor (for example NotePad, TextEdit) and examine it.

::: {.callout-exercise #ex_weather}

{{< fa solid anchor >}}

Based on this table, what was the temperature in degrees C on 2026-07-02 at 12:00:00?

::: {.callout-answer collapse="true"}
17.8
:::

:::


### Reading CSV Files

The R function `read_csv` reads a CSV file into R. 

If I wanted to read the file `weather_data.csv`, the code I would use would be as follows. I can view the data frame with the same `gt` function we used previously.

```{r read_weather_data}
weather <- read_csv("data/weather_data.csv")
gt(weather)
```

::: {.callout-exercise #ex_public_health}

{{< fa solid flask >}}

### Reading UK Public Health Monitoring Data

The UK government provide some public health data for England (other UK countries have data available elsewhere) on their data portal, here - https://ukhsa-dashboard.data.gov.uk/. The data is available for download in CSV format.

Choose any of the datasets with measurements provided over time which are available on this page and download the CSV formatted data to your computer. You can find these by clicking on one of the graphs then choosing "Download", then "CSV format". Move this file into the `Practical_1/data` folder of your course 

In a new Quarto report, load this table into R and then view the resulting table with `gt()`. These tables are mostly long, so we can use the `head()` function to just visualise the first ten lines:

For example:

```{r gt_head}
rhinovirus <- read_csv("data/ukhsa-chart-download-rhinovirus.csv")

gt(head(rhinovirus))
```

Don't forget to write a line or two describing what you did - for example, which dataset you have selected.

You will need this section at the top of your Quarto document:
```{r load_tidyverse2}
library(tidyverse)
library(gt)
```

:::
## Plotting in R {#sec-plots}

[↑ top](#)

### Initiating a Plot
We will now learn to draw some basic graphs in R, we will learn much more about this over future weeks.

We start a new, empty plot using the function `ggplot`. We pass this function our dataframe variable, to tell it that this is the data we are going to plot.

In this example, I'll plot data from the `weather` data frame.

```{r load_weather, echo=FALSE}
gt(weather)
```

First, I read the table and initiate a blank canvas on which to put my plot.

```{r initiate_plot}
weather <- read_csv("data/weather_data.csv")

ggplot(weather)
```


### Adding axes
We use another function, `aes`, as an argument to the `ggplot` function,  telling R how various attributes of the data map to the "aesthetics" of the plot.

For example, here we choose which data we will represent with the *x* and *y* axes.

```{r axes_plot}
ggplot(weather, aes(
  x = observation_time,
  y = temperature_c
))
```

### Adding Data

Now we can add some data to our graph. In this plotting package, `ggplot`, we add additional things to our plot using a `+` symbol.

Here, we add a line representing the data using the `geom_line` function.

```{r line_plot}
ggplot(weather, aes(
  x = observation_time,
  y = temperature_c
)) +
  geom_line()
```

### Adding extra features

We can add some additional features here to make the plot look nicer. For example, here I will change the colour of the line to red, add *x* and *y* axis labels and add a title.

```{r colour_line_plot}
ggplot(weather, aes(
  x = observation_time,
  y = temperature_c
)) +
  geom_line(color = "red") +
  xlab("Date") +
  ylab("Temp (C)") +
  ggtitle("Temperature by Date")
```

::: {.callout-exercise #ex_public_health_plot}

{{< fa solid flask >}}

Following on from the UKHSA public health data you downloaded above and loaded into R, in the same Quarto document:

* Plot a line graph, where the x axis is date (contained in the `date` column), and the *y* axis is the metric measured in the table (from the `metric_value`) column.
* Change the colour of the line and add appropriate axis labels - you can go back to the UKHSA data website to see exactly what is being measured.
:::

## Summary {#sec-summary_0104}

[↑ top](#)

* We can combine multiple variables of the same type into a **vector** using the function `c`.
* Tabular data can be represented in a **data frame**, generated using the function `tibble`.
* Data is commonly made available in comma-seperated variable (CSV) format
* The function `read_csv` is used to read external data in CSV format into an R data frame.
* Plots are generated from data using the function `ggplot`.
* We can add axes and other aesthetics to the plot using the funcition `aes`.

{{< include ../includes/git_push.qmd >}}
