---
format: html
practical: Practical_3_Probability
doc-id: "0307"
toc-depth: 4
---
# Probability Problem Sheets

These solutions use primarily simulation to solve the problems, rather than calculating exact mathematical solutions.

```{r setup_0307}
library(tidyverse)
library(Runuran)
library(DescTools)
```

## Problem Sheet 1
We will use 10,000 simulations, unless otherwise specified.

```{r set_sims}
n_sim <- 10000
```

### Question 1

In a natural habitat, grey wolves, foxes and coyotes co-exist. The current population
is estimated to be 60 foxes, 30 coyotes and 10 wolves. If each animal is equally likely
to be sighted first, find the probability of:
a. a coyote being sighted first
b. a wolf being sighted first
c. a fox being sighted second after either a coyote or a wolf has been sighted and
then removed from the habitat, so they can be fitted with a tracking device
(Consider: what is the impact of the first sighting on what you see next?)


```{r q1}
# Make a population with 60 foxes, 30 coyotes and 10 wolves
animals <- c(
  rep("fox", 60),
  rep("coyote", 30),
  rep("wolf", 10)
)

# Sample 10000 animals, with replacement
samps <- sample(
  x = animals,
  size = n_sim,
  replace = TRUE
)
```

#### a
Probability that a coyote is sighted first.

```{r q1a}
# Count how many samples are coyotes
n_coy <- sum(samps == "coyote")

# What proportion of the total samples is this?
ans <- n_coy / length(samps)

print(ans)
```

The probability a coyote is sighted first is `r round(ans, 4)` 

#### b
Probability that a wolf is sighted first.

```{r q1b}
# Count how many samples are wolves
n_wolves <- sum(samps == "wolf")

# What proportion of the total samples is this?
ans <- n_wolves / length(samps)
print(ans)
```
The probability a wolf is sighted first is `r round(ans, 4)` 

#### c
Probability of a fox being sighted second after either a coyote or a wolf has been sighted and then removed from the habitat, so they can be fitted with a tracking device.

```{r q1c}
times_w_c_first <- 0
times_f_second <- 0

first_animal <- c()
second_animal <- c()
for (i in seq_len(10000)) {
  # Sample 2 animals from the habitat
  # and remove each after sampling
  samps <- sample(
    x = animals,
    size = 2,
    replace = FALSE
  )
  # Track who was caught first
  first_animal <- c(first_animal, samps[1])

  # Track who was caught second
  second_animal <- c(second_animal, samps[2])
}

# Vector of all the times wolf or coyote was
# first
w_c_first <- first_animal %in% c("wolf", "coyote")

# Vector of all the times fox was second
fox_second <- second_animal == "fox"

# Count how many times a wolf or coyote was
# first
w_c_first_count <- sum(w_c_first)

# Count how many times a wolf or coyote was
# first and a fox was second
w_c_first_and_fox_second_count <- sum(w_c_first & fox_second)

# Calculate the proportion of the times that
# a wolf or coyote was first that a fox was
# also second
ans <- w_c_first_and_fox_second_count / w_c_first_count
print(ans)
```
The probability of a fox being sighted second after either a coyote or a wolf has been sighted and
then removed from the habitat `r round(ans, 4)` 


### Question 2
A transmitter emits the signals “0” or “1” in the ratio 1:2. Two independent receivers,
labelled R1 and R2 are set up. If, unknown to the users, receiver R1 is incorrect in 2%
of the cases and receiver R2 is incorrect in 3% of the cases, find

a. the probability that R1 reports that the transmitted signal is “0”
b. the probability that R2 reports that the transmitted signal is “1”
c. following an incoming signal there is some confusion, because R1 reports
signal “0” and R2 reports signal “1”. Which of the two receivers is most likely
to be correct?

```{r q2}
# Run the transmitter 10,000 times, with a 2/3 probability of success
# where success == 1
transmitter <- rbinom(n = n_sim, size = 1, prob = 2 / 3)

# Run R1 10,000 times, with a 98% probability of success
# 1 - correct
# 0 - incorrect
R1_correct <- rbinom(n = n_sim, size=1, prob = 0.98)

# Run R2 10,000 times, with a 97% probability of success
# 1 - correct
# 0 - incorrect
R2_correct <- rbinom(n = n_sim, size=1, prob = 0.97)

```

#### a

The probability that R1 reports that the transmitted signal is “0”

```{r q2a}
# R1 reports 0 when either the signal is 0
# and it is correct or the signal is 1
# and it is incorrect
zero_and_correct <- transmitter == 0 & R1_correct == 1
one_and_incorrect <- transmitter == 1 & R1_correct == 0

# Find times when either of these are true
# | = OR
R1_output_of_zero <- zero_and_correct | one_and_incorrect

# Count how many times either were true
count_R1_output_of_zero <- sum(R1_output_of_zero)

# Convert to a proportion
ans_a <- count_R1_output_of_zero / n_sim
print(ans_a)
```
The probability that R1 reports that the transmitted signal is “0” is `r round(ans_a, 4)`

#### b

The probability that R2 reports that the transmitted signal is “1”

```{r q2b}
# R2 reports 1 when either the signal is 1
# and it is correct or the signal is 0
# and it is incorrect
one_and_correct <- transmitter == 1 & R2_correct == 1
zero_and_incorrect <- transmitter == 0 & R2_correct == 0

# Find times when either of these are true
R2_output_of_one <- one_and_correct | zero_and_incorrect

# Count how many times either were true
count_R2_output_of_one <- sum(R2_output_of_one)
ans_b <- count_R2_output_of_one / n_sim

print(ans_b)
```
The probability that R2 reports that the transmitted signal is “1” is `r round(ans_b, 4)`

#### c

c. following an incoming signal there is some confusion, because R1 reports signal “0” and R2 reports signal “1”. Which of the two receivers is most likely to be correct?

```{r q2c}
# Find when R1 reports 0 and R2 reports 1
R1_0_R2_1 <- (R1_output_of_zero == 1) & (R2_output_of_one == 1)

# Count how many of these times R1 reports 0 and is correct
R1_reports_zero_correctly <- sum(R1_correct == 1 & R1_0_R2_1 == 1)

# Count how many of these times R1 reports 0 and is correct
R2_reports_one_correctly <- sum(R2_correct == 1 & R1_0_R2_1 == 1)

# Calculate the proportion of times R1 is correct
R1_correct_prop <- R1_reports_zero_correctly / sum(R1_0_R2_1 == 1)
R2_correct_prop <- R2_reports_one_correctly / sum(R1_0_R2_1 == 1)


print (paste("R1 = ", R1_correct_prop))
print (paste("R2 = ", R2_correct_prop))
```

Given that R1 has reported 0 and R2 has reported 1, R2 is more likely to be correct, with a probability of `r round(R2_correct_prop, 4)`, compared to `r round(R1_correct_prop, 4)` for R1.

### Question 3

a. Of a batch of 22 imported oak saplings, 12 are healthy and 10 have hidden oak
processionary moth eggs. A tree health officer is aware that there may be an
issue and chooses 5 trees at random to inspect in detail. As there are slightly
more healthy trees than infested trees, we might predict that the batch will
consist of 3 healthy trees and 2 infested trees. How many ways are there to
select 3 healthy trees and 2 infested trees from the 22 saplings?

b. What is the probability that a sample of five trees, chosen at random from the
batch of 22 imported trees, consists of three healthy and two infested trees?

c. Suppose instead that all 5 of the trees selected are healthy (and so the
infestation remains undetected). How many ways are there that the sample
could have been chosen at random from the batch to give this result?

d. What is the probability the sample of five, chosen at random from the batch
consists only of healthy trees?

::: {.callout-note #note-choose}
For a and c, this simulation will usually give approximately the right number of combinations, but it may not find every answer.
:::

#### a

How many ways are there to select 3 healthy trees and 2 infested trees from the 22 saplings?
```{r q3a}
# Create a population of 12 healthy trees and
# 10 infested trees, all with unique names
healthy <- paste("healthy", seq(1, 12))
eggs <- paste("eggs", seq(1, 10))

# Take three healthy trees 10,000 times
# Keep track of all combinations encountered
# Sort by name as order doesn't matter
healthy_samps <- c()
for (i in seq_len(n_sim)) {
  samp <- sample(x = healthy, size = 3, replace = FALSE)
  samp_s <- paste(samp[order(samp)], collapse = " ")
  healthy_samps <- c(healthy_samps, samp_s)
}

# Take two infested trees 10,000 times
# Keep track of all combinations encountered
# Sort by name as order doesn't matter
egg_samps <- c()
for (i in seq_len(n_sim)) {
  samp <- sample(x = eggs, size = 2, replace = FALSE)
  samp_s <- paste(samp[order(samp)], collapse = " ")
  egg_samps <- c(egg_samps, samp_s)
}

# Find all the unique combinations encountered
unique_healthy <- length(unique(healthy_samps))
unique_eggs <- length(unique(egg_samps))

# Multiply these together to get the total
# number of combinations of 3 healthy and 2
# infested
total_count <- unique_healthy * unique_eggs
print (total_count)
```
There are `r total_count` ways to select 3 healthy and 2 infested.

Note that this only works because the outcome is fairly likely - for improbable outcomes more simulations would be needed to encounter every possible outcome.

#### b

What is the probability that a sample of five trees, chosen at random from the
batch of 22 imported trees, consists of three healthy and two infested trees?

```{r q3b}
# Make a batch of 12 healthy and 10 infested trees
healthy <- rep("healthy", 12)
eggs <- rep("eggs", 10)
combined <- c(healthy, eggs)

# Sample 5 trees 10,000 times and track how many are
# healthy each time.
healthy_counts <- c()
for (i in seq_len(n_sim)) {
  samp <- sample(x = combined, size = 5, replace = FALSE)
  n_healthy <- sum(samp == 'healthy')
  healthy_counts <- c(healthy_counts, n_healthy)
}
# Count how many times 3 were healthy
total_three_healthy <- sum(healthy_counts == 3)

# Divide by the number of simulations
prop <- total_three_healthy / n_sim
print(prop)
```

There is a probability of `r round(prop, 4)` that 3 trees are healthy and two are infested.

#### c
Suppose instead that all 5 of the trees selected are healthy (and so the infestation remains undetected). How many ways are there that the sample could have been chosen at random from the batch to give this result?

```{r q3c}
# Make a sample of 12 healthy trees, with unique
# names
healthy <- paste("healthy", seq(1, 12))

# Take five healthy trees 10,000 times
# Keep track of all combinations encountered
# Sort by name as order doesn't matter
healthy_samps <- c()
for (i in seq_len(10000)) {
  samp <- sample(x = healthy, size = 5, replace = FALSE)
  samp_s <- paste(samp[order(samp)], collapse = " ")
  healthy_samps <- c(healthy_samps, samp_s)
}

unique_healthy <- unique(healthy_samps)
n_unique_healthy <- length(unique_healthy)
print(n_unique_healthy)
```
There are `r n_unique_healthy` ways to select 5 healthy trees.

Note that this only works because the outcome is fairly likely - for improbable outcomes more simulations would be needed to encounter every possible outcome.

#### d
What is the probability the sample of five, chosen at random from the batch consists only of healthy trees?

```{r q3d}
# Use the sample from part b - 5 trees sampled 10,000
# times, tracked how many are healthy each time.

# Count how many times 5 were healthy
total_five_healthy <- sum(healthy_counts == 5)

# Divide by the number of simulations
prop <- total_five_healthy / n_sim
print(prop)
```
There is a probability of `r round(prop, 4)` that all 5 trees are healthy.

### Question 4

A spam filter on a mail server scrutinises messages for the appearance of key words
that make it likely that the message is spam.

The filter works as follows:
• it is assumed that evidence for spam is that the subject message of the email
contains the sentence “check this out”.

According to some historical data we know that:
• 40% of emails are spam
• 1% of spam emails have “check this out” in the subject line
• 0.4% of non-spam emails have “check this out” in the subject line.
Find the probability that an email is spam given that it has the phrase “check this
out” in the subject

```{r q4}
# Simulate 10,000 emails with a 40% chance of being spam
# 1 = spam
# 0 = not spam
is_spam <- rbinom(100000, size = 1, prob = 0.4)

# Count how many of these are and are not spam
n_spam <- sum(is_spam == 1)
n_not_spam <- 100000 - n_spam

# For the results which are spam, simulate a 1% chance
# of saying CTO
is_cto_spam <- rbinom(n_spam, size=1, prob = 0.01)

# For the results which are not spam, simulate a 0.4% chance
# of saying CTO
is_cto_not_spam <- rbinom(n_not_spam, size=1, prob = 0.004)

# Calculate the total number of simulations that are CTO
cto = sum(is_cto_spam, is_cto_not_spam)

# Calculate the proportion of CTO simulations which are spam
prob_spam_cto <- sum(is_cto_spam) / cto

print (prob_spam_cto)
```
The probability that an email is spam given that it has the phrase “check this out” in the subject is `r round(prob_spam_cto, 4)`

Note that I used a higher number of simulations here because so few emails have "check this out" overall.

### Question 5

The number of particles emitted from a radioactive substance follows a Poisson distribution with parameter λ=4 particles per minute. Find the probability that:

a. three particles are emitted in one minute
b. at least one particle is emitted in a minute
c. three particles are emitted in 2 minutes

```{r q5}
# Simulate a Poisson distributed data set with a lambda of 4
particles_per_min <- rpois(n_sim, 4)
```

#### a
Find the probability that three particles are emitted in one minute

```{r q5a}
# How many of the simulated data points have a count of exactly 3
ans_a <- sum(particles_per_min == 3) / n_sim
print(ans_a)
```

The probability that 3 particles are emitted in one minute is `r round(ans_a, 4)`.

#### b

Find the probability that at least one particle is emitted in a minute
```{r q5b}
# How many of the simulated data points have a count of at least 1
ans_b <- sum(particles_per_min >= 1) / n_sim
print(ans_b)
```

The probability that at least one particle is emitted in a minute is `r round(ans_b, 4)`.

#### c

Find the probability that 3 particles are emitted in 2 minutes

```{r q5c}
# Simulate a Poisson distributed data set with a lambda of 8
particles_per_2min <- rpois(n_sim, 8)
# How many of the simulated data points have a count of exactly 3
ans_c <- sum(particles_per_2min == 3) / n_sim
print (ans_c)
```

The probability that 3 particles are emitted in 2 minutes is `r round(ans_c, 4)`.

### Question 6

a. The survival rate of a specific disease is forty percent. In a certain day eight
patients are admitted into hospital with that disease. What are the chances
that three will survive?

b. A new drug is discovered that can cure the disease but can only be
administered in a way that has a success rate of reaching the affected area of
0.6. There usually are four doses of the drug administered. Find the chances
of:
  i. three doses successfully administered
  ii. no doses successfully administered
  iii. at least one dose successfully administered
  iv. at most two doses successfully administered

#### a
The survival rate of a specific disease is forty percent. In a certain day eight patients are admitted into hospital with that disease. What are the chances that three will survive?
```{r q6a}
# Sample 8 individuals from a binomial distribution 
# 10,000 times with a 40% probability of survival
survival <- rbinom(n_sim, 8, 0.4)

# Calculate the number of simulated values which are
# equal to 3, convert to a proportion of 10,000
n_three <- sum(survival == 3)
prop_three <- n_three / n_sim

print(prop_three)
```

The probability that three will survive is `r round(prop_three, 4)`.

A new drug is discovered that can cure the disease but can only be administered in a way that has a success rate of reaching the affected area of 0.6.
```{r q6b}
# Simulate giving four doses of the drug 10,000
# times with a 0.6 probability of success
success <- rbinom(10000, 4, 0.6)
```

##### i
Three doses successfully administered
```{r q6bi}
# Number of times there were exactly 3 successes
ans_i <- sum(success == 3) / 10000
print (ans_i)
```
The probability that three doses were successfully adminstered is `r round(ans_i, 4)`

##### ii
No doses successfully administered
```{r q6bii}
# Number of times there were exactly 0 successes
ans_ii <- sum(success == 0) / 10000
print (ans_ii)
```
The probability that no doses were successfully adminstered is `r round(ans_ii, 4)`

##### iii
At least one dose successfully administered
```{r q6biii}
# Number of times at least one success (i.e. not zero successes)
ans_iii <- sum(success != 0) / 10000
print (ans_iii)
```
The probability that at least one dose was successfully administered is `r round(ans_iii, 4)`

##### iv
At most two doses successfully administered
```{r q6biv}
# Number of times 2 or less successes
ans_iv <- sum(success <= 2) / 10000
print (ans_iv)
```
The probability that at most two doses were successfully administered is `r round(ans_iv, 4)`

### Question 7

A scientist randomly selects dolphins until she finds one that has been tagged with a
tracking device, so that she can collect the data on the pod’s movements. The
probability that she finds the dolphin with the tracking device is 0.2.

a. What is the probability that the scientist must select exactly 4 dolphins before
she finds one with a tracking device?

b. What is the probability that she must select more than 6 dolphins before she
finds one with a tracking device?

```{r q7}
# Simulate a geometric distribution with 10,000 data points, with a 
# 20% probability of success in each trial.
dolphins <- rgeom(n_sim, 0.2)
```

#### a
What is the probability that the scientist must select exactly 4 dolphins before she finds one with a tracking device?

```{r q7a}
# Probability there are 3 failures before the first success
ans_a <- sum(dolphins == 3) / n_sim
print (ans_a)
```

There is a `r round(ans_a, 4)` probability that she must select exactly 4 dolphins to find one with a tracking device.

#### b
b. What is the probability that she must select more than 6 dolphins before she finds one with a tracking device?

```{r q7b}
# Probability there are at least 6 failures before the first success
ans_b <- sum(dolphins >= 6) / n_sim
print (ans_b)
```

There is a `r round(ans_b, 6)` probability that she will have to select more than 6 dolphins before finding one with a tracking device.

### Question 8

The number of results we get per minute from a simple computational model is a Poisson random variable with mean 3.

a. Find the probability that no results come in a given 1 minute period.

b. Find the probability that at least two results will come in a given two minute period.

```{r q8}
# Simulate a Poisson distribution with lambda 3
results_per_min <- rpois(n_sim, 3)
```

#### a
Find the probability that no results come in a given 1 minute period.

```{r q8a}
# Probability of a count of exactly 0
ans_a <- sum(results_per_min == 0) / n_sim
print (ans_a)
```
The probability no results come in a given 1 minute period is `r round(ans_a, 4)`.

#### b
The probability that at least two results will come in a given two minute period.
```{r q8b}
# Simulate a Poisson distribution with lambda 6
results_per_2min <- rpois(10000, 6)
# Probability of a count of at least 2
ans_b <- sum(results_per_2min >= 2) / n_sim
print (ans_b)
```
The probability at least two results will come in a given 2 minute period is `r round(ans_b, 4)`.

## Problem Sheet 2

### Question 1

On average, a certain spectrophotometer part lasts 10 years. The length of time that the
part lasts is exponentially distributed.
a) What is the probability that the spectrophotometer part lasts more than 7 years?
b) On the average, how long would 5 of these parts last if they were used one after
another?

```{r qq1}
# Rate per year if it lasts 10 years on average
rate = 1 / 10

# Simulate an exponential distribution with this
# rate and 10,000 observations
distri <- rexp(n_sim, rate=rate)
```

#### a
What is the probability that the spectrophotometer part lasts more than 7 years?
```{r qq1a}
# Proportion of intervals > 7
ans_a <- sum(distri > 7) / n_sim
print (ans_a)
```
There is an `r round(ans_a, 4)` probability the part lasts more than 7 years.

#### b
On the average, how long would 5 of these parts last if they were used one after another?
```{r qq1b}
# Average time based on simulation
average_time <- mean(distri)

# Multiply by 5
total_time <- average_time * 5

print (total_time)
```
On average, 5 parts would last `r round(total_time, 4)` years.

### Question 2

A drug is produced in a chamber with temperature X, where X is a random variable that follows a normal distribution with μ = 21°C and σ = 5°C. If X<25°C, the production cost per unit is £20, whereas if X≥25°C the production cost per unit is £30. Find the expected production cost per unit for this drug.

```{r qq2}
# Simulate a normal distribution where μ = 21°C and σ = 5°C
drug_dist <- rnorm(n_sim, 21, 5)

# Count how many times X is < 25
cold <- sum(drug_dist < 25)

# Count how many times X is >= 25
hot <- sum(drug_dist >= 25)

# Calculate the total cost for cold production
cold_cost <- cold * 20

# Calculate the total cost for hot production
hot_cost <- hot * 30

# Calculate the total cost for all production
total_cost <- cold_cost + hot_cost

# Convert to a proportion of the number of units
ans_2 <- total_cost / n_sim

print (ans_2)
```
The expected production cost per unit for this drug is £`r round(ans_2, 2)` per unit.

### Question 3

A box contains 3 green and 5 red balls. If we pick balls and then replace them, find:
a) The probability that the third time we pick a ball, we get a red ball for the second time.
b) The probability that the 4th time we pick a ball, a red ball appears for the first time
c) The expected number of times we pick a ball until we get a red ball for the second time.


#### a
The probability that the third time we pick a ball, we get a red ball for the second time.
```{r qq3a}
# Draw two balls 10,000 times and count how many
# are red
first_two <- rbinom(n_sim, size = 2, prob = 5/8)

# Draw one ball 10,000 times and count how many
# are red
third <- rbinom(n_sim, size = 1, prob = 5/8)

# How many times is either ball 1 or ball 2 red
# but not both and ball 3 is red
count <- sum(first_two == 1 & third == 1)

# As a proportion
prop <- count / n_sim

print (prop)
```
The probability that the third time we pick a ball, we get a red ball for the second time is `r round(prop, 4)`.

#### b
The probability that the 4th time we pick a ball, a red ball appears for the first time
```{r qq3b}
# Draw three balls 10,000 times and count how many
# are red
first_three <- rbinom(n_sim, size = 3, prob = 5/8)

# Draw one ball 10,000 times and count how many
# are red
fourth <- rbinom(n_sim, size = 1, prob = 5/8)

# How many times are none of the first three balls
# red but ball 4 is red
count <- sum(first_three == 0 & fourth == 1)

# As a proportion
prop <- count / n_sim

print (prop)
```
The probability that the 4th time we pick a ball, a red ball appears for the first time is `r round(prop, 4)`.

#### c
The expected number of times we pick a ball until we get a red ball for the second time.

```{r qq3c}
# Simulate the number of green balls picked before the first red
# twice
ball_1 <- rgeom(n_sim, 5/8)
ball_2 <- rgeom(n_sim, 5/8)

# Add them together - add two for the two red balls then picked
combined_balls <- ball_1 + ball_2 + 2

# The mean of these values is the expectation
expected <- mean(combined_balls)

print (expected)
```
The expected number of times we pick a ball until we get a red ball for the second time is `r round(expected, 4)`.


### Question 4
For a group of 100 students, find:
a) The expected number of days of the year that are birthdays of exactly 3 people.
b) The expected number of distinct birthdays


```{r qq4}
prob_bday <- 1 / 365

bdays_per_day <- rbinom(1000000, 100, prob_bday)
```

```{r qq4a}
prob_3_bdays <- sum(bdays_per_day == 3) / 1000000

n_days <- 365 * prob_3_bdays

print (n_days)
```

```{r qq4b}
prop_distinct <- sum(bdays_per_day != 0) / 1000000
n_distinct <- 365 * prop_distinct

print (n_distinct)
```

### Question 5

::: {.callout-note #note-non_standard_dist}
Answering this question using simulation depends on the bonus material in @sec-non-standard.
:::

A continuous random variable X has the following probability density function:
$$
f(x) =
\begin{cases}
\displaystyle\frac{4x(9-x^2)}{81}, & \text{if } 0 \leq x \leq 3 \\[6pt]
0, & \text{otherwise}
\end{cases}
$$

a. Verify that f(x) is a valid probability density function.
b. Find (a) the mean of f(x), (b) the mode of f(x) and (c) the median of f(x)


#### a

Verify that f(x) is a valid probability density function.

```{r qq5a}
# Calculate the area under the curve
area <- integrate(function (x) {((4 * x) * (9 - x^2)) / 81},
                  lower=0,
                  upper=3)
print(area)
```
The area under the curve is `r area$value` so the PDF is valid.

#### b

Find (a) the mean of f(x), (b) the mode of f(x) and (c) the median of f(x)

```{r qq5b}
# Make a random number generator using this function
rand_gen <- pinv.new(
  pdf = function(x) {((4 * x) * (9 - x^2)) / 81},
  lb = 0,
  ub = 3
)

# Generate 10,000 random points
simu <- ur(rand_gen, n_sim)

# Calcualte the mean and median of the points
mean_simu <- mean(simu)
median_simu <- median(simu)

# Mean
print(mean_simu)

# Median
print (median_simu)
```
The simulated mean is `r round(mean_simu, 4)`.

The simulated median is `r round(median_simu, 4)`.

The mode is trickier, as R doesn't by default calculate the mode for continuous data. It's easiest to do this visually based on a histogram.

```{r mode_cont}
# Convert to a data frame
simu_df <- tibble(simulated_data = simu)

# Plot a histogram
ggplot(simu_df, aes(x = simulated_data)) + 
  geom_histogram(bins=50)
```
The peak is apparent on this plot at about 1.7.

### Question 6

::: {.callout-note #note-non_standard_dist}
Answering this question using simulation depends on the bonus material in @sec-non-standard.
:::

### a
Let X be a continuous random variable with probability density function:

$$
f(x) =
\begin{cases}
4x^3, & \text{if } 0 < x \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$

Find

$$
P\left(X \leq \frac{2}{3} \mid X > \frac{1}{3}\right).
$$
```{r qq6a}
# Create a random number generator for this distribution
rand_gen <- pinv.new(
  pdf = function(x) {4 * x^3},
  lb = 0,
  ub = 1
)

# Generate 10,000 random points
simu <- ur(rand_gen, n_sim)

# Convert to a data frame
rand_df <- tibble(points = simu)

# Extract the points where x > 1/3
# within this interval
rand_df_2_3 <- rand_df |> 
  filter(points <= (2/3),
         points > 1/3)

# Extract the points where x > 1/3
rand_df_1_3 <- rand_df |> 
  filter(points > (1/3))

# Calculate the proportion of points within the interval
prop_between <-
  nrow(rand_df_2_3) / nrow(rand_df_1_3)

print (prop_between)
```

### b

Let \(X\) be a continuous random variable with probability density function:

$$
f(x) =
\begin{cases}
x^2\left(2x+\frac{3}{2}\right), & \text{if } 0 < x \leq 1 \\
0, & \text{otherwise}
\end{cases}
$$

If

$$
Y = \frac{2}{X} + 3,
$$

find

$$
\operatorname{Var}(Y).
$$
```{r qq6b}
# Create a random number generator for this function
rand_gen <- pinv.new(
  pdf = function(x) {x^2 * (2*x + (3/2))},
  lb = 0,
  ub = 1
)

# Generate 100,000 random points
# This one isn't very stable so use a bigger number
simu <- ur(rand_gen, 100000)

# Calculate the Y value for each point and store it.
simu_y <- c()
for (sim_x in simu){
  y = (2 / sim_x) + 3
  simu_y <- c(simu_y, y)
}

# Find the variance of Y
var_y <- var(simu_y)

print (var_y)
```
The variance of Y is `r round(var_y, 4)`.

The simulated estimate of the variance may differ noticeably from the exact answer of approximately 1.97, because values of X close to zero produce very large values of Y, making the sample variance relatively unstable.

### Question 8

A polar bear is hunting for food. From its current position, there are three different
routes. The first one brings it back to the same point after 1 hours of walking. The
second route brings it back to the same point after 6 hours of walking. The last route
leads to the food source after 2 hours of walking. Assuming that the probability the
bear chooses route 1 is 0.25, route 2 is 0.25 and route 3 is 0.5, what is the expected
time it will take for the bear gets some food?

::: {.callout-note #note-while_loops}

This solution uses conditional statements **if** and **else**, along with a  **while** loop - a loop which runs over and over again until something becomes true. We haven't covered these yet, but we will in Practicals 4 and 6. 
:::

```{r qq8}
# Length of each route
route1 <- 1
route2 <- 6
route3 <- 2

# Probability of each route
prob1 <- 0.25
prob2 <- 0.25
prob3 <- 0.5

# Currently, the bear has no food
food <- FALSE

# Vector to track the time taken in each simulation
total_times <- c()

for (i in seq_len(n_sim)){
  total_time <- 0
  food = FALSE
  # Keep running these steps until he finds food
  while (!food) {
    # Randomly, based on the probabilities above, choose a route
    which_route <- sample(seq_len(3), 1, prob = c(prob1, prob2, prob3))
    
    # If it is route 1, add 1 hour to the total time
    if (which_route == 1){
      total_time <- total_time + route1
    }
    # If it is route 2, add 6 hours to the total time
    else if (which_route == 2){
      total_time <- total_time + route2
    }
    # If it is route 3, add 2 hours to the total time
    # then stop the loop because he has found food.
    else if (which_route == 3){
      total_time <- total_time + route3
      food = TRUE
    }
  }
  # Store the total time taken
  total_times <- c(total_times, total_time)
}
# Calculate the mean of the time taken over the 10,000 simulations
mean_total <- mean(total_times)
print (mean_total)
```
The mean time for the polar bear to find food is `r round(mean_total, 4)`.