---
format: html
practical: Practical_3_Probability
doc-id: "0304"
---
# Part 4: Simulation exercises {#sec-using_other_distributions}

::: {.callout-note .partmenu #parts-0303}

## Sections
- @sec-distribution_1
- @sec-distribution_2

:::

:::{.callout-tip .objectives #objectives-0303}

## Learning Objectives

By the end of this section, you will have practiced:

* Simulating data based on any probability distribution implemented in R.
* Comparing observed data with theoretical probability distributions.
* Using simulation to investigate whether observed data are consistent with a distribution.

:::

```{r setup_0304, include=FALSE}
library(tidyverse)
library(gt)
library(patchwork)
set.seed(42)
```


Using the skills we've built so far, we can now simulate data from any probability distribution implemented in R, whether to compare with observed data, test a hypothesis or model a process.

### Distribution 1 {#sec-distribution_1}

The data in `data/neurons.csv` is from [this paper](https://doi.org/10.3389/fncom.2019.00069), published by Koepcke et al. in Frontiers in Computational Neuroscience in 2019. 

These authors investigated whether a single neuron can rapidly detect changes in a sensory stimulus by analysing the time between successive action potentials (interspike intervals). The analysis was performed using multiple neurons.

The data in data/neurons.csv come from a single neuron and contain a series of waiting times between successive spikes, measured in seconds.

::: {.callout-exercise #ex-neurons}

Imagine that you are a researcher and you notice that this neuron seems to fire very regularly. In particular, it never seems to pause for very long between spikes. By estimating the probability distribution of the waiting times and using simulation, investigate whether the longest observed wait is unusually short.

In your `distributions.qmd` Quarto file:

* Read the CSV file containing the data.
* Plot a **histogram** with an appropriate number of bins showing how the data is distributed.
* Based on the shape of the histogram and the biological process being measured, decide which probability distribution might provide an appropriate model for the waiting times.
* Find the longest wait time in the observed data.
* Using the appropriate function from the table in @sec-other_distributions, generate 1000 random datasets with the same number of measurements from the distribution you have selected, generating parameters from your data as needed. Record the longest wait time.
* Assuming the observed data follows the same distribution as the simulated data, plot a histogram of your simulated maxima and estimate if the maximum observed wait time is unexpectedly short. 
* Calculate the percentage of your simulated samples in which the maximum wait time is **less** than or equal to the maximum observed in the study.

::: {.callout-note .extension #bon-pexp}

### Extension Exercise: Cumulative probability
* Using the appropriate cumulative probability function, calculate the theoretical probability that a single wait time will be less than or equal to your observed maximum.

:::

::: {.callout-answer #ex-neurons_ans}

[Read the file containing the data.]{.blue-text}

```{r read_data_neurons, message=FALSE}
neuron_df <- read_csv("data/neurons.csv")
```

[Plot a **histogram** with an appropriate number of bins showing how the data is distributed.]{.blue-text}

```{r plot_data_neurons}
ggplot(
  neuron_df,
  aes(x = Inter_Spike_Interval)
) +
  geom_histogram(bins = 50)
```
[Based on the shape of the histogram and the biological process being measured, decide which probability distribution might provide an appropriate model for the waiting times.]{.blue-text}

The distribution of waiting times between spikes looks approximately exponential.

If instead we were to count the number of spikes in a fixed interval, the simplest model would be to assume they are Poisson distributed.

* We are counting how many times an event - a spike - occurs.
* Counts are made over a known interval.
* Spikes occur independently (we assume).
* Spikes occur at a constant average rate throughout the interval (again, we assume).
* The chance of multiple spikes in a very small interval is negligible.
* There is no fixed upper limit to the count - in theory any number of spikes could occur.

If spike occurrences can be modelled as a Poisson process, the waiting times between successive spikes follow an exponential distribution.

[Find the longest wait time in the observed data.]{.blue-text}

```{r max_neurons}
max_observed <- max(pull(neuron_df, Inter_Spike_Interval))
print(max_observed)
```
[Using the appropriate function from the table in @sec-other_distributions, generate 1000 random datasets with the same number of measurements from the distribution you have selected, generating parameters from your data as needed. Record the longest wait time.]{.blue-text}

```{r calc_rate_neurons}
sample_mean_wait_time <- mean(pull(neuron_df, Inter_Spike_Interval))
n_observations <- nrow(neuron_df)
lambda <- 1 / sample_mean_wait_time
```

The mean wait time in our dataset is `r round(sample_mean_wait_time, 4)`, so the rate for our exponential distribution is 1/`r round(sample_mean_wait_time, 4)`.

```{r simulate_neurons}
maxis <- c()
for (i in seq_len(1000)) {
  expo <- rexp(
    n = n_observations,
    rate = lambda
  )
  maxi <- max(expo)
  maxis <- c(maxis, maxi)
}
maxi_df <- tibble(Maximum_Wait_Time = maxis)
```

[Assuming the observed data follows the same distribution as the simulated data, plot a histogram of your simulated maxima and estimate if the maximum observed wait time is unexpectedly long or short.]{.blue-text}

```{r plot_neurons_max}
ggplot(
  maxi_df,
  aes(x = Maximum_Wait_Time)
) +
  geom_histogram(bins = 50) +
  geom_vline(xintercept = max_observed, color = "red")
```
The observed maximum waiting time is towards the lower end of the simulated maxima, but still falls within the range expected under the fitted exponential model.

[Calculate the percentage of your simulated samples in which the maximum wait time is **less** than or equal to the maximum observed in the study.]{.blue-text}

```{r calc_neurons_max}
filt_maxi <- maxi_df |>
  filter(Maximum_Wait_Time <= max_observed)

prop <- nrow(filt_maxi) / nrow(maxi_df)
print(prop)
```


::: {.callout-note .extension #bon-pexp_ans}
### Extension Exercise: Cumulative probability - Answer

[Using the appropriate cumulative probability function, calculate the theoretical probability that a single wait time will be less than or equal to your observed maximum.]{.blue-text}

```{r theor_neurons_one}
prob_one_wt <- pexp(max_observed, rate = lambda)
print(prob_one_wt)
```

[Calculate the theoretical probability that **all** `r nrow(neuron_df)` of your observations would be less than this value.]{.blue-text}

```{r theor_neurons_all}
prob_all_wt <- prob_one_wt^n_observations
print(prob_all_wt)
```
:::

Based on these results, the observed maximum waiting time is shorter than many of the simulated maxima. However, it is still well within the range expected from an exponential distribution, so there is no strong evidence that its behaviour is unusual, assuming our wait times are exponentially distributed.

:::
:::

## Distribution 2 {#sec-distribution_2}

::: {.callout-note .extension #bon-weibull}

### Extension Exercise: Distribution 2 

Pancreatic islets of Langerhans are clusters of endocrine cells in the pancreas. Individual pancreatic islets vary in size - measurements from human pancreatic islet isolations show a distribution of islet diameters.

In [this paper](https://doi.org/10.3727/096368909X476968), published in Cell Transplantation in 2009, Buchwald et al. analysed the size composition of human pancreatic islet isolations. The authors had data from more than 200 human islet isolations and looked at the distribution of islet diameters. They found lots of relatively small islets and progressively fewer large ones.

They found that the distribution of individual islet diameters could be described by a Weibull probability distribution with shape parameter κ=1.5 and scale parameter λ=105.

The Weibull probability distribution is available in R using the functions `rweibull()`, `dweibull()`, `pweibull()` and `qweibull`. 

The parameters for `rweibull` are:
* n - number of points to simulate
* shape - the shape parameter of the distribution
* scale - the scale parameter of the distribution

In your `distributions.qmd` Quarto file:
* Plot a density-scaled histogram of the diameters of 10,000 simulated islets.
* Overlay a theoretical Weibull distribution.
* During manual assessment, only islets with a diameter of **at least 50 μm** are included in the standard counting procedure. Given that an islet is included in the count, estimate the probability that its diameter is **between 100 and 150 μm**.

::: {.callout-answer collapse="true" #ex-weibull_ans}

[Plot a density-scaled histogram of the diameters of 10,000 simulated islets.]{.blue-text}

[Overlay a theoretical Weibull distribution.]{.blue-text}

```{r weibull_ans}
# Simulate the islets
islet_diams <- rweibull(
  n = 10000,
  shape = 1.5,
  scale = 105
)
# Put them in a data frame
islet_diams_df <- tibble(Islet_Diameter = islet_diams)

# Plot the results as a histogram with an overlaid
# Weibull distribution
ggplot(islet_diams_df, aes(x = Islet_Diameter)) +
  geom_histogram(aes(y = after_stat(density)),
    bins = 50,
    fill = "hotpink"
  ) +
  stat_function(
    fun = dweibull,
    args = list(
      shape = 1.5,
      scale = 105
    )
  )
```
[During manual assessment, only islets with a diameter of **at least 50 μm** are included in the standard counting procedure. Given that an islet is included in the count, estimate the probability that its diameter is **between 100 and 150 μm**.]{.blue-text}

```{r prob_islets}
# Filter the data frame to keep only islets >= 50 um
islets_over_50 <- islet_diams_df |> filter(Islet_Diameter >= 50)

# Count how many of these there are
n_islets_over_50 <- nrow(islets_over_50)

# Filter the data frame to keep only islets 100-150 um
islets_100_150 <- islet_diams_df |> filter(
  Islet_Diameter >= 100,
  Islet_Diameter <= 150
)
# Count how many of these there are
n_islets_100_150 <- nrow(islets_100_150)

# Calculate the proportion of islets >= 50 um which are 100 to 150 um
prop_counted_islets_100_150 <- n_islets_100_150 / n_islets_over_50
print(prop_counted_islets_100_150)
```

:::
:::

For this part of the practical, the main files you have edited are:

* `distributions.qmd`

These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}

