---
format: html
practical: Practical_3_Probability
doc-id: "0302"
author: "Katy Brown"
---


```{r setup_0301, include=FALSE}
library(tidyverse)
library(gt)
library(cowsay)
library(DescTools)
library(patchwork)
set.seed(42)
```

# Part 2: Using simulation to understand probability  {#sec-p2_simulation}

::: {.callout-note .partmenu #parts-0301}

## Sections
- @sec-simulation
- @sec-how_simulation
- @sec-unequal_probabilities
- @sec-repeating_simulations
- @sec-binomial
- @sec-inbuilt_functions
:::

:::{.callout-tip .objectives #objectives-0301}

## Learning Objectives
By the end of this part of the practical you will be able to

- Approximate the outcome of stochastic processes using simulation
- Use `sample()` to simulate random events with equal or unequal probabilities.
- Use a `for` loop to repeat a simulation.
- Calculate binomial probabilities using `dbinom()` and `pbinom()`.
- Generate random values from a binomial distribution using `rbinom()`.
- Compare an empirical distribution obtained by simulation with a theoretical probability mass function.
:::

## Introduction to simulation {#sec-simulation}

Many biological processes involve an element of chance and therefore involve probability. In R, we can use simulation to generate data that approximates a random processes. These simulated data allow us to investigate probability, test hypotheses and better understand the behaviour of biological systems.

If you are studying a stochastic process, the exact probability of an event can **sometimes** be calculated mathematically. In these cases, simulation provides a useful way to **check** your calculations. In other cases, an exact solution may be difficult or impossible to obtain, and simulation allows you to **estimate** the answer instead.

## How to simulate data {#sec-how_simulation}

First, let's perform a tiny simulation, of a coin toss.

We are using the function [`sample()`](../../R_reference.qmd#sec-prob_sim_glos), with the arguments `x` - the elements from which to sample, and `size`, the number of samples to take. By default, R assumes that every element is equally probable. 

```{r toss_one_coin}
coin_toss <- sample(c("Heads", "Tails"), size = 1)
print(coin_toss)
```

We can use a for loop to toss multiple coins.

```{r toss_five_coins}
for (i in seq_len(5)) {
  coin_toss <- sample(c("Heads", "Tails"),
                      size=1)
  print(coin_toss)
}
```

We can also do this directly with the sample function by changing `size`. We need to add the argument `replace = TRUE` to tell R to replace the samples - otherwise after two samples, the coin has no remaining sides!

```{r toss_five_coins_sample}
coin_tosses <- sample(c("Heads", "Tails"),
                      size=5,
                      replace = TRUE)
print(coin_tosses)
```

To count how many times a specific value occurs in a vector, you can use the operator `==` and the `sum()` function:

```{r count_coins}
n_successes <- sum(coin_tosses == "Heads")
print(n_successes)
```

The coin showed Heads `r n_successes` times.

## Unequal probabilities {#sec-unequal_probabilities}

So far, we've assumed a "fair coin" - with 50% probability of each outcome. However, we can change this by giving an additional argument to the `sample()` function.

This argument is `prob` - a vector of **probability weights** corresponding to the elements of the original vector.

So, if we want the coin to land on heads 70% of the time, we would use:

```{r coin_biased}
coin_tosses <- sample(c("Heads", "Tails"),
  size = 5,
  replace = TRUE,
  prob = c(0.7, 0.3)
)
print(coin_tosses)
```

If we had a vector `c('Badger', 'Mushroom', 'Snake')` and we want to sample 100 times from this vector, with replacement, [selecting `Badger` 95% of the time, `Mushroom` 4% of the time and `Snake` 1% of the time](https://www.youtube.com/watch?v=EIyixC9NsLI), we would use this function as follows:

```{r badger_badger_mushroom}
bm_vec <- c("Badger", "Mushroom", "Snake")
bm_probs <- c(0.95, 0.04, 0.01)

my_sample <- sample(bm_vec,
  size = 100,
  replace = TRUE,
  prob = bm_probs
)
print(my_sample)
```


::: {.callout-exercise #ex-sample_binom}

A squirrel has a 30% success rate at stealing seeds from a bird feeder. She tries 150 times.

In your `squirrels.qmd` Quarto file, using the `sample()` function, take 150 samples with a 30% probability of a successful heist. Calculate the total number of successful heists and store it as a variable named `n_successes`.

Have a squirrel announce the results of your simulation (the number of times she was successful) using the function `say()` and the argument `by="squirrel"`.

::: {.callout-answer collapse="true" #ex_sample_binom_ans}
```{r sample_binom_ans}
is_success <- sample(c("SUCCESS", "FAILURE"),
  size = 150,
  replace = TRUE,
  prob = c(0.3, 0.7)
)
n_successes <- sum(is_success == "SUCCESS")

say(paste("I was successful", n_successes, "times!"), by = "squirrel")
```
:::
:::

## Repeating simulations {#sec-repeating_simulations}

If you run this simulation several times, you'll see that the result often changes. 

We can formalise this using a for loop, running the simulation a fixed number of times.

Let's say we have 75 squirrels, each making 150 attempts, with the same fixed probability of success of 30%.

```{r for_loop_squirrels}
combined_successes <- c()
for (i in seq_len(75)) {
  is_success <- sample(c("SUCCESS", "FAILURE"),
    size = 150,
    replace = TRUE,
    prob = c(0.3, 0.7)
  )
  n_successes <- sum(is_success == "SUCCESS")
  combined_successes <- c(combined_successes, n_successes)
}
print(combined_successes)
```

We now know how many successful heists each squirrel had, out of 150.

Let's visualise how many squirrels have each number of successful heists - i.e. how many of our 75 squirrels stole seeds once, twice, three times etc.

First, we need to convert our vector into a data frame.

```{r vec_data_frame_squirrels}
df_squirrels_75 <- tibble("N_Successful_Heists" = combined_successes)
gt(head(df_squirrels_75))
```

The we can create a plot showing how many squirrels had each number of successful heists.

The [`geom_bar()`](../../R_reference.qmd#sec-plotting_glos) function in `ggplot` counts how many times each value occurs in a column of a data frame and displays the result as a bar chart.

`geom_bar()` just only needs the x axis to be specified - in this case the `N_Successful_Heists` column, as the y axis will be calculated.


```{r plot_squirrels_75}
ggplot(
  df_squirrels_75,
  aes(
    x = N_Successful_Heists)
) +
  geom_bar()
```
```{r store_75, include=FALSE}
# store the plot to display later
squirrels_75 <- ggplot(
  df_squirrels_75,
  aes(
    x = N_Successful_Heists)
) +
  geom_bar()
```

You can see that a distribution is emerging here - the squirrels are likely to be successful around 40 to 55 times out of 150. 

This is called an **empirical distribution** because it is based on simulated observations rather than a mathematical formula.

## The binomial distribution {#sec-binomial}

As you may have recognised, this approximates a binomial distribution:

* We have a fixed number of trials - 150.
* We have a binary outcome - success or failure.
* Our trials are independent.
* The probability of success is constant across trials.

The number of successes announced by the squirrel above is the random variable - the number of successes, *k* out of 150 trials.

Increasing the number of simulations reduces the random variation in our estimate, bringing it closer to the theoretical distribution. Therefore, if we run our simulation more times, our empirical approximation should get closer to a theoretical binomial distribution.


::: {.callout-exercise #ex-big_binom}
In your `squirrels.qmd` Quarto document, adapt the code to perform the simulation with 750 squirrels, and again plot the results.

::: {.callout-answer collapse="true" #ex_big_binom_ans}
```{r more_squirrels}
# 750 squirrels
# Make an empty vector to store the results
combined_successes <- c()

# For each of 750 squirrels
for (i in seq_len(750)) {
  # Have the squirrel attempt to steal seeds 150 times with
  # a 30% probability of success
  is_success <- sample(c("SUCCESS", "FAILURE"),
    size = 150,
    replace = TRUE,
    prob = c(0.3, 0.7)
  )
  # Count the number of successes
  n_successes <- sum(is_success == "SUCCESS")
  # Store the result in our vector
  combined_successes <- c(combined_successes, n_successes)
}

# Convert the results into a data frame
combined_successes_df <- tibble(n_successes = combined_successes)

df_squirrels_750 <- tibble("N_Successful_Heists" = combined_successes)

# Plot the results
ggplot(
  df_squirrels_750,
  aes(
    x = N_Successful_Heists)
) +
  geom_bar()
```
```{r store_750, include=FALSE}
# store the plot to display later
squirrels_750 <- ggplot(
  df_squirrels_750,
  aes(
    x = N_Successful_Heists)
) +
  geom_bar()
```

:::
:::

Your new plot should look even closer to a binomial distribution.

## Using probability functions in R

As you learned in the lectures, the **probability mass function** (PMF) for the binomial distribution is:

$$
P(X = k) =
\binom{n}{k}
p^k
(1-p)^{n-k}
$$


This is implemented in R with the function `dbinom()`.

The arguments for this function are:

* `x`: calculate the number of this number of successes - *k* in the equation, 
* `size`: number of trials, *n* in the equation, in our case 150 attempts per squirrel
* `prob`: probability of success in each trial - *p* in the equation, the probability of squirrel success, 0.3, in our case.

Let's estimate the probability that a squirrel succeeds 48 times.


```{r pmf_once}
n <- 150
k <- 48
p <- 0.3

prob_X_equals_k <- dbinom(k, n, p)
print(prob_X_equals_k)
```

We would expect this result around 6% of the time.

Let's compare this with the observed result from our simulation. Using filter, we can view the row showing how many squirrels succeeded 48 times.

```{r observed_squirrels}
n_successes <- pull(df_squirrels_750, N_Successful_Heists)
prop_48 <- sum(n_successes == 48)
print (prop_48)
```

In my version of the simulation, `r prop_48` out of 750 squirrels succeeded 48/150 times, your results might vary.

```{r calc_prop}
print(prop_48 / 750)
```
This is around `r round((prop_48 / 750) * 100)`% of the squirrels - not far off the estimate using the PMF.

::: {.callout-note .extension #bon-theoretical_binom}

### Extension content: Plotting the theoretical distribution

Using a for loop, we can try this for every value of `k` - i.e. every number of successes, from 0 to 150.

```{r pmf_probs}
n <- 150
p <- 0.3
probs <- c()
for (k in seq(0, n, 1)) {
  prob_X_equals_k <- dbinom(k, n, p)
  probs <- c(probs, prob_X_equals_k)
}
probs
```

As above, let's convert our results into a data frame. We know our results correspond to values of k from 0 to 150, so we can add these as a column.

```{r pmf_df}
df_probs <- tibble(
  N_Successful_Heists = seq(0, n, 1),
  Probability = probs
)
```

And now we can make a bar plot of these probabilities.

This time, instead of using [`geom_bar()`](../../R_reference.qmd#sec-plotting_glos), which calculates the number of occurances of each value in a column, we already have already calculated the heights we want for the bars in our bar chart - the probability of each number of successes. Therefore, we can use the function [`geom_col()`](../../R_reference.qmd#sec-plotting_glos) which draws a bar chart based directly on the values in a data frame column.

`geom_col()` just  needs the x and y axes to be specified - in this case `N_Successful_Heists` on the x axis and `Probability` on the y axis. as the y axis will be calculated.


```{r pmf_bar_plot}
ggplot(
  df_probs,
  aes(
    x = N_Successful_Heists,
    y = Probability
  )
) +
  geom_col()
```

```{r store_pmf_bar_plot, include=FALSE}
squirrels_pmf <- ggplot(
  df_probs,
  aes(
    x = N_Successful_Heists,
    y = Probability
  )
) +
  geom_col()
```

This plot shows how the data would be distributed under a perfect, theoretical binomial distribution.

:::

Now let's compare the plots based on our two simulations with the theoretical distribution. In order to compare them easily, I have set all three to have the same axis limits and given each a title.

```{r all_plots_binom, fig.height=12, echo=FALSE}
plot_75_xl <- squirrels_75 +
  coord_cartesian(xlim = c(25, 65)) +
  ggtitle("75 Squirrels")

plot_750_xl <- squirrels_750 +
  coord_cartesian(xlim = c(25, 65)) +
  ggtitle("750 Squirrels")

plot_pmf_xl <- squirrels_pmf +
  coord_cartesian(xlim = c(25, 65)) +
  ggtitle("PMF")

plot_75_xl / plot_750_xl / plot_pmf_xl
```

As you can see, we've successfully recreated a binomial distribution using random sampling. With even more squirrels, our approximation would be even closer.

## Inbuilt functions {#sec-inbuilt_functions}

R is often used for simulation, so there are inbuilt functions to help us.

Up to this point, we have simulated each individual attempt and then counted the number of successes. Because this type of experiment is so common, R provides a function that generates the number of successes directly.

With this binomially distributed data, we could have avoided most of the above (sorry) by using the function [`rbinom()`](../../R_reference.qmd#sec-prob_sim_glos), which automatically draws samples from a binomial distribution.

This function takes three arguments:

* `n`: number of observations, in our case e.g. 750 squirrels
* `size`: number of trials, in our case 150 attempts per squirrel
* `prob`: probability of success, in our case 0.3 (30% chance of stealing seeds successfully)


```{r use_rbinom}
count_squirrels <- rbinom(n = 750, size = 150, prob = 0.3)
print(count_squirrels)
```

This gives us our counts per squirrel, just like we calculated in @ex-big_binom above, and we can plot it in the same way.

```{r plot_rbinom}
rb_df <- tibble(N_Successful_Heists = count_squirrels)
ggplot(rb_df, aes(x = N_Successful_Heists)) +
  geom_bar()
```

We have already used the function [`dbinom()`](../../R_reference.qmd#sec-prob_sim_glos), which gives us the probability of a specific number of successes, given specified values of k, n and p.

Another function, [`pbinom()`](../../R_reference.qmd#sec-prob_sim_glos) calculates the probability that there will be **no more than** a specific number of successes. This is known as the cumulative probability. The arguments are similar to `dbinom`:

* `q`: calculate the probability of no more than this number of successes.
* `size`: number of trials - *n* in the equation, the number of squirrels, 150, in our case.
* `prob`: probability of success in each trial - *p* in the equation, the probability of squirrel success, 0.3, in our case.

For example, to calculate the probability that a squirrel will succeed no more than 48 times:

```{r pbinom}
prob_up_to_48 <- pbinom(48, 150, 0.3)
print(prob_up_to_48)
```

We can verify this with our `count_squirrels` vector generated with `rbinom`.

```{r check_pbinom}
print(sum(count_squirrels <= 48) / length(count_squirrels))
```

::: {.callout-exercise #ex-rbinom_flowers}

A botanist studies a plant species in which each flower has a 60% probability of having been pollinated. She needs 125 pollinated flowers for her experiment, but she can't tell which flowers are pollinated until she gets back to the lab. She picks 200 flowers.

Using simulation, estimate the probability that she has collected enough pollinated flowers for her experiment.

Please solve this problem using the `rbinom()` function in your `prac3_exercises.qmd` Quarto file.

::: {.callout-note #note-sumif}
To count how many numbers in a numeric vector, <code>[vec]{.orange-text}</code>, are less than or equal to a certain value, <code>[val]{.blue-text}</code> you can use <code>sum([vec]{.orange-text} <= [val]{.blue-text})</code>. This syntax also works with other operators (@sec-filter), e.g. `<`, `>`, `>=`, `==`, `!=`.
:::

::: {.callout-answer collapse="true" #ex_rbinom_flowers_ans}


```{r rbinom_flowers_ans1}
n_sims <- 1000
n_flowers_per_sim <- 200

# Simulate 1000 datasets of 200 flowers with
# a 60% chanec of being pollinated
flowers <- rbinom(
  n = n_sims,
  size = n_flowers_per_sim,
  prob = 0.6
)

# How many times were at least 125 flowers
# pollinated?
times_successful <- sum(flowers >= 125)

# What proportion of the attempts is this?
prop_successful <- times_successful / n_sims
print(prop_successful)
```

Based on this simulation, there is an estimated `r round(prop_successful*100)`% probability she has 125 pollinated flowers.

We can check this using `pbinom()`.

```{r rbinom_flowers_ans2}
# Probability that less than 125 flowers are
# pollinated
prob_under_125 <- pbinom(124, 200, 0.6)

# Probability that more than 125 flowers are
# pollinated
prob_125_or_over <- 1 - prob_under_125

print(prob_125_or_over)
```

`r round(prob_125_or_over*100)`% - similar to the simulated value. Running more simulations would make the estiamte more accurate.

:::
:::

::: {.callout-note .extension  #bon-binomial}

### Extension exercise

Estimate the smallest number of flowers the botanist would need to collect to have at least a 95% probability of obtaining 125 pollinated flowers.

::: {.callout-answer collapse="true` #bon-binomial_ans}
Option 1: Using simulation

```{r rbinom_flowers_ans4}
n_sims <- 1000
props <- c()
# We know it's more than 200 - let's try to up to 250
# Run the simulation for each of these numbers of
# flowers
for (i in seq(200, 250, 1)) {
  flowers <- rbinom(
    n = n_sims,
    size = i,
    prob = 0.6
  )
  # Calculate the number and proportion of times
  # she was successful
  # Store only this value, not the whole vector
  times_successful <- sum(flowers >= 125)
  prop_successful <- times_successful / n_sims
  props <- c(props, prop_successful)
}

# Convert these into a data frame
prop_df <- tibble(
  npicked = seq(200, 250, 1),
  psuccess = props
)

# Find the rows where she succeeded at least
# 95% of the time
prop_df_filt <- prop_df |>
  filter(psuccess >= 0.95)

gt(prop_df_filt)
```

In my version of the simulation, she needs to pick `r pull(prop_df_filt, npicked)[1]` flowers.

Option 2: Using `pbinom`

```{r rbinom_flowers_ans5}
probs <- c()
# We know it's more than 200 - let's try to up to 250
# Determine the probability for each of these numbers
# of flowers
for (i in seq(200, 250, 1)) {
  prob_under_125 <- pbinom(124, i, 0.6)
  prob_125_or_over <- 1 - prob_under_125
  probs <- c(probs, prob_125_or_over)
}

# Convert these into a data frame
prob_df <- tibble(
  npicked = seq(200, 250, 1),
  psuccess = probs
)

# Find the rows where she succeeded at least
# 95% of the time
prob_df_filt <- prob_df |>
  filter(psuccess >= 0.95)

gt(prob_df_filt)
```

She needs to pick `r pull(prop_df_filt, npicked)[1]` flowers.

:::
:::


## Review {#sec-summary_0301}

[↑ top](#)

As you have seen above, simulation is a very useful tool for approximating a probability distribution and answering questions about stochastic processes.

::: {.callout-note #note-summary_0301}

### Programming summary

* `sample()` generates random samples with equal or unequal probabilities.
* `for` loops can be used to repeat a simulation many times.
* `dbinom()` calculates probabilities from a binomial distribution.
* `pbinom()` calculates cumulative probabilities from a binomial distribution.
* `rbinom()` generates random values from a binomial distribution

:::


For this part of the practical, the main files you have edited are:

* `squirrels.qmd`

These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}
