---
format: html
practical: Practical_3_Probability
doc-id: "0305"
---

```{r setup_0305, include=FALSE}
library(tidyverse)
library(gt)
set.seed(42)
```

# Part 5: Conditional Probability {#sec-p5_bayes}

::: {.callout-note .partmenu #parts-0305}

## Sections
- @sec-dataset_cc
- @sec-hpv_data
- @sec-variation_with_age_cc
- @sec-cc_as_outcome [(Bonus)]{.bonus-text}
- @sec-summary_0305

:::

::: {.callout-tip .objectives #objectives-0305}

## Learning objectives

By the end of this part of the practical you will be able to:

- Calculate conditional probabilities from observed data in R.
- Calculate and interpret the sensitivity and specificity of a biological marker.
- Use simulation to investigate conditional probabilities.
- Explain how the prevalence of an outcome affects the interpretation of a marker or test result.

:::

## UK cervical screening data {#sec-dataset_cc .core}

[**Core**]{.core-text}

[↑ top](#)

In the UK, cervical screening is routinely offered regularly to any individual with a cervix aged 25 to 64. This screening tests for high-risk strains of **human papillomavirus** (HPV). Members of this group of viruses can damage cells in the cervix and cause cervical cancer. Where tests are positive for **high-risk HPV (hrHPV)**, the samples are further tested for abnormal cells and, if abnormal cells are found, additional tests for pre-cancerous cells and cervical cancer are performed.

For the purposes of this practical, we will treat hrHPV status as a marker associated with abnormal cervical cytology, and later explore how this relates to the probability of cervical cancer.

The following data are from [this publication](https://doi.org/10.1038/sj.bjc.6603210), published in 2006 in the British Journal of Cancer (note that this is before routine vaccination against HPV was introduced in the UK).

In this study 24,510 people in the UK were screened for hrHPV and cytology was performed in all cases to detect abnormal cells. In this practical we will consider 21,843 of these people - by excluding people who had borderline or mild abnormal cytology.

## HPV data analysis {#sec-hpv_data .core}

[**Core**]{.core-text}

[↑ top](#)

Of these 21,843, 12.1% of people tested positive for hrHPV. 2.1% of people had moderately or severely abnormal cytology, the remainder were normal.

The number of people with and without hrHPV and with and without abnormal cytology are provided in CSV format in `data/cm_all_ages.csv`. 

:::{.callout-exercise #ex-confusion_matrix_all_ages .core}

### Calculate conditional probabilities

[**Core**]{.core-text}

In your `hpv.qmd` Quarto file, read `data/cm_all_ages.csv` into a data frame and view the contents with `gt()`.

For this first part of the exercise you do not need to use simulation, you can directly calculate the values from the data frame.

Calculate:
* The proportion of people with hrHPV who have abnormal cells. Store this as `prop_abnormal_given_hpv`
* The proportion without hrHPV who have abnormal cells. Store this as `prop_abnormal_given_no_hpv`
* The proportion of people with abnormal cells who have hrHPV - the sensitivity of hrHPV as a marker. Store this as `prop_hpv_given_abnormal`.
* The proportion of people who have normal cells and don't have hrHPV - the specificity of hrHPV as a marker. Store this as `prop_no_hpv_given_normal`.

:::{.callout-note #note-filter_pull}

Remember that you can use the `pull()` function to extract a column from a data frame and the `filter()` function to filter rows based on specific criteria.

:::

:::{.callout-answer collapse="true" #ex-confusion_matrix_all_ages_ans}

[In your `hpv.qmd` Quarto file, read `data/cm_all_ages.csv` into a data frame and view the contents with `gt()`.]{.blue-text}

```{r confusion_matrix_all_ages, message=FALSE}
hpv_df <- read_csv("data/cm_all_ages.csv")
gt(hpv_df)
```

Extract the true positives, true negatives, false positives, false negatives:

* True positives - have HPV and have abnormal cells
* True negatives - don't have HPV and have normal cells
* False positives - have HPV and have normal cells
* False negatives - don't have HPV and have abnormal cells

```{r extract_values_hrhpv}
tp <- hpv_df |>
  filter(
    HPV_Status == "HPV+",
    Cytology == "Abnormal"
  ) |>
  pull(N_individuals)
tn <- hpv_df |>
  filter(
    HPV_Status == "HPV-",
    Cytology == "Normal"
  ) |>
  pull(N_individuals)
fp <- hpv_df |>
  filter(
    HPV_Status == "HPV+",
    Cytology == "Normal"
  ) |>
  pull(N_individuals)
fn <- hpv_df |>
  filter(
    HPV_Status == "HPV-",
    Cytology == "Abnormal"
  ) |>
  pull(N_individuals)
```

[Calculate the proportion of people with hrHPV who have abnormal cells.]{.blue-text}

```{r prob_hrHPV_ab, message=FALSE}
# Count how many people have hrHPV - positives
hpv_count <- tp + fp

# Calculate the proportion of people with hrHPV who
# also have abnormal cells - true positives / positives
prop_abnormal_given_hpv <- tp / hpv_count
print(prop_abnormal_given_hpv)
```
`r round(prop_abnormal_given_hpv*100, 2)`% of people with hrHPV have abnormal cells.

[Calculate the proportion of people without hrHPV who have abnormal cells.]{.blue-text}

```{r prob_nhrHPV_ab, message=FALSE}
# Count how many people don't have hrHPV - negatives
hpv_m_count <- tn + fn

# Calculate the proportion of people who don't have hrHPV who
# do have abnormal cells - false_negatives / negatives
prop_abnormal_given_no_hpv <- fn / hpv_m_count
print(prop_abnormal_given_no_hpv)
```
`r round(prop_abnormal_given_no_hpv*100, 2)`% of people without hrHPV have abnormal cells.


[Calculate the proportion of people with abnormal cells who have hrHPV - the sensitivity of the hrHPV as a marker.]{.blue-text}

$$
\text{Sensitivity} =
\frac{\text{True positives}}
{\text{True positives} + \text{False negatives}}
$$


```{r prob_ab_hrHPV, message=FALSE}
# Count how many people have abnormal cells -
# true positives and false negatives
ab_count <- tp + fn

# Calculate the proportion of people who have
# abnormal cells who also have hpv
prop_hpv_given_abnormal <- tp / ab_count
print(prop_hpv_given_abnormal)
```

The sensitivity of HPV status as a test for abnormal cells is `r round(prop_hpv_given_abnormal * 100, 2)`%.

[Calculate the proportion of people who have normal cells and don't have hrHPV - the specificity of hrHPV as a marker.]{.blue-text}

$$
\text{Specificity} =
\frac{\text{True negatives}}
{\text{True negatives} + \text{False positives}}
$$

```{r prob_n_nhrHPV, message=FALSE}
# Count how many people have normal cells -
# true negatives and false positives
n_count <- tn + fp

# Calculate the proportion of people who have
# normal cells who don't have hpv
prop_no_hpv_given_normal <- tn / n_count
print(prop_no_hpv_given_normal)
```

The specificity of HPV status as a test for abnormal cells is `r round(prop_no_hpv_given_normal * 100, 2)`%.

:::
:::

## Variation with age {#sec-variation_with_age_cc .rec}

[**Recommended**]{.rec-text}

[↑ top](#)

So far, we've calculated proportions using the whole sample of 21,843 individuals. However, the prevalence of moderate/severe abnormal cytology varies considerably with age.

In this sample the authors described:

* **5.08%** prevalence of moderate/severe abnormal cytology in individuals aged 20-29
* **1.81%** prevalence of moderate/severe abnormal cytology in individuals aged 30-49
* **0.47%** prevalence of moderate/severe abnormal cytology individuals aged 50-64

In the following series of exercises, we will use simulation to deduce the probability that someone with hrHPV has moderate/severe abnormal cytology.

:::{.callout-exercise #ex-simulation_by_age .rec}

### Simulate different age groups

[**Recommended**]{.rec-text}

In your `hpv.qmd` Quarto file:

* For each of these age groups, simulate a sample of 10,000 individuals with the appropriate prevalence of moderate/severe cytology. You can use a binomial distribution with the function `rbinom()` where `n_sim` is 10000, `size` is 1 and `prob` is the proportion of people with moderate/severe morphology from the table above.

* For each age range, count how many samples in the simulation have moderate/severe cytology and how many do not.

::: {.callout-answer collapse="true" #ex-simulation_by_age_ans1}

[For each of these age groups, simulate a sample of 10,000 individuals with the appropriate prevalence of moderate/severe cytology. You can use a binomial distribution with the function `rbinom()` where `n_sim` is 10000, `size` is 1 and `prob` is the proportion of people with moderate/severe morphology from the table above.]{.blue-text}

```{r simulation_by_age_ans1}
cyt_statuses_20_29 <- rbinom(10000, size = 1, prob = 0.0508)
cyt_statuses_30_49 <- rbinom(10000, size = 1, prob = 0.0181)
cyt_statuses_50_64 <- rbinom(10000, size = 1, prob = 0.0047)
```

[For each age range, count how many samples in the simulation have moderate/severe cytology and how many do not.]{.blue-text}

```{r simulation_by_age_ans2}
n_mod_sev_20_29 <- sum(cyt_statuses_20_29)
n_ok_20_29 <- 10000 - n_mod_sev_20_29
n_mod_sev_30_49 <- sum(cyt_statuses_30_49)
n_ok_30_49 <- 10000 - n_mod_sev_30_49
n_mod_sev_50_64 <- sum(cyt_statuses_50_64)
n_ok_50_64 <- 10000 - n_mod_sev_50_64
```

:::
:::

Now we know how many people have normal and abnormal cells. The next step is to simulate which of the people with and without abnormal cells are hrHPV positive.

In @ex-confusion_matrix_all_ages we already calculated the probability of an individual having hrHPV, given that they have abnormal cells, and stored this in the variable `prop_hpv_given_abnormal`. 

We didn't calculate the probability of an individual having hrHPV given that they have normal cells, but we did calculate the probability of **not** having hrHPV given that they have normal cells - this is stored as `prop_no_hpv_given_normal`. We can therefore subtract this from 1 to get the probability of having HPV given that they have normal cells.

::: {.callout-note .copycode #copy-prop_hpv_given_normal}

### Include this code 

Include this code in your `hpv.qmd` Quarto file.

```{r prop_hpv_given_normal}
prop_hpv_given_normal <- 1 - prop_no_hpv_given_normal
```

:::

:::{.callout-exercise #ex-simulation_by_age2 .rec}

### Investigate the effect of prevalence

[**Recommended**]{.rec-text}

In your `hpv.qmd` Quarto file:

* For each age range, for the number of individuals in the simulation with moderate / severe cytology, calculate the expected number with HPV, again using `rbinom()` but this time with `size` as the number of individuals with moderate/severe cytology and `prob` as `prop_hpv_given_abnormal`.

* For each age range, for the number of individuals in the simulation without moderate / severe cytology, calculate the expected number with HPV, again using `rbinom()` but this time with `size` as the number of individuals without moderate/severe cytology and `prob` as `prop_hpv_given_normal`.

* For each age range, find the total number of people with HPV by adding together the number of people with HPV and moderate / severe cytology and the number of people with HPV and not moderate / severe cytology.

* Store the number of people with HPV, the number of people with abnormal cells and the number of people with abnormal cells and HPV in a data frame, with one row for each age range. You may find it easiest to create a one-row data frame for each age range and combine them using `bind_rows()`.

::: {.callout-answer collapse="true" #ex-simulation_by_age2_ans}

[For each age range, for the number of individuals in the simulation with moderate / severe cytology, calculate the expected number with HPV, again using `rbinom()` but this time with `size` as the number of individuals with moderate/severe cytology and `prob` as `prop_hpv_given_abnormal`.]{.blue-text}

```{r simulation_by_age_ans3}
hpv_count_ab_20_29 <- sum(rbinom(n_mod_sev_20_29, size = 1, prob = prop_hpv_given_abnormal))
hpv_count_ab_30_49 <- sum(rbinom(n_mod_sev_30_49, size = 1, prob = prop_hpv_given_abnormal))
hpv_count_ab_50_64 <- sum(rbinom(n_mod_sev_50_64, size = 1, prob = prop_hpv_given_abnormal))
```

[For each age range, for the number of individuals in the simulation without moderate / severe cytology, calculate the expected number with HPV, again using `rbinom()` but this time with `size` as the number of individuals without moderate/severe cytology and `prob` as `prop_hpv_given_normal`.]{.blue-text}

```{r simulation_by_age_ans4}
hpv_count_norm_20_29 <- sum(rbinom(n_ok_20_29, size = 1, prob = prop_hpv_given_normal))
hpv_count_norm_30_49 <- sum(rbinom(n_ok_30_49, size = 1, prob = prop_hpv_given_normal))
hpv_count_norm_50_64 <- sum(rbinom(n_ok_50_64, size = 1, prob = prop_hpv_given_normal))
```

[For each age range, find the total number of people HPV by adding together the number of people with HPV and moderate / severe cytology and the number of people with HPV and not moderate / severe cytology.]{.blue-text}

```{r simulation_by_age_ans5}
n_hpv_20_29 <- hpv_count_ab_20_29 + hpv_count_norm_20_29
n_hpv_30_49 <- hpv_count_ab_30_49 + hpv_count_norm_30_49
n_hpv_50_64 <- hpv_count_ab_50_64 + hpv_count_norm_50_64
```

[Store the number of people with HPV, the number of people with abnormal cells and the number of people with abnormal cells and HPV in a data frame, with one row for each age range.]{.blue-text}

```{r simulation_by_age_ans6}
row_20_29 <- tibble(
  age_range = "20-29",
  n_hpv = n_hpv_20_29,
  n_abnormal_cells = n_mod_sev_20_29,
  n_hpv_and_abnormal_cells = hpv_count_ab_20_29
)

row_30_49 <- tibble(
  age_range = "30-49",
  n_hpv = n_hpv_30_49,
  n_abnormal_cells = n_mod_sev_30_49,
  n_hpv_and_abnormal_cells = hpv_count_ab_30_49
)

row_50_64 <- tibble(
  age_range = "50-64",
  n_hpv = n_hpv_50_64,
  n_abnormal_cells = n_mod_sev_50_64,
  n_hpv_and_abnormal_cells = hpv_count_ab_50_64
)

results <- bind_rows(row_20_29, row_30_49, row_50_64)
gt(results)
```

:::
:::

Now, based on these data, we can deduce the probability that someone with hrHPV will go on to be diagnosed with moderate/severe abnormal cytology for each age group. 

::: {.callout-exercise #ex-simulation_by_age3 .rec}

### Count HPV positive individuals

[**Recommended**]{.rec-text}

In your `hpv.qmd` Quarto file:

* Use your simulated data to estimate the probability that an individual in each group with hrHPV has moderate-or-worse abnormal cytology. 

* Why does the probability that an individual in each group with hrHPV has moderate-or-worse abnormal cytology differ so much, even though the sensitivity and specificity of hrHPV as a marker were kept constant?

::: {.callout-answer collapse="true" #ex-simulation_by_age3_ans}

[Use your simulated data to estimate the probability that an individual in each group with hrHPV has moderate-or-worse abnormal cytology. ]{.blue-text}

```{r simulation_by_age_ans7}
results <- results |> mutate(
  Abnormal_given_hrHPV = n_hpv_and_abnormal_cells / n_hpv
)

gt(results)
```

* `r round(results |> filter(age_range == "20-29") |> pull(Abnormal_given_hrHPV)*100, 2)`% of people aged 20-29 with HPV have abnormal cells.
* `r round(results |> filter(age_range == "30-49") |> pull(Abnormal_given_hrHPV)*100, 2)`% of people aged 30-49 with HPV have abnormal cells.
* `r round(results |> filter(age_range == "50-64") |> pull(Abnormal_given_hrHPV)*100, 2)`% of people aged 50-64 with HPV have abnormal cells.


[Why does the probability that an individual in each group with hrHPV has moderate-or-worse abnormal cytology differ so much, even though the sensitivity and specificity of hrHPV as a marker were kept constant?]{.blue-text}

Because the prevalence of abnormal cells changes so much between the groups. The sensitivity of the test tells us the probability the person will test positive if they have the disease and the specificity tells us the probability they will test negative if they don't have the disease. The ratio of people who do and don't have the disease differs between the three samples.

:::
:::

## Cervical cancer as an outcome {#sec-cc_as_outcome .fold .bonus}

[**Bonus**]{.bonus-text}

We can extend this simulation to consider cervical cancer as an outcome, rather than just abnormal cells.

Based on [this paper](https://doi.org/10.1111/cyt.12259) published in Cytopathology in 2016 and also based on UK individuals, the positive predictive value of moderate or severely abnormal cells for cervical cancer is:

* **1.5%** for people aged 20-29
* **3.4%** for people aged 30-49
* **5.1%** for people aged 50-64

(Note that these are estimates, as this paper uses slightly different age ranges)

This means that, based on this sample, for example, 1.5% of people aged 20-29 with at least moderately abnormal cells will then be diagnosed with cervical cancer. In this study no-one of any age with normal cells was diagnosed with cervical cancer.

::: {.callout-exercise #ex-cc_outcome .bonus}

### Cervical cancer as an outcome

[**Bonus**]{.bonus-text}

Extend your simulation to calculate the probability that a 24 year old who is HPV positive and has a cervix would be diagnosed with cervical cancer on further testing (assuming prevalence is constant across the 20-29 age group).

::: {.callout-answer collapse="true" #ex-cc_ans }

```{r cc_ans}
results_cc <- tibble()
prop_cc_given_abnormal <- 0.015
prop_abnormal <- 0.0508
n_sim <- 10000

# Simulate a population of people with normal and
# abnormal cells at the population rate for age 20-29
# 1 = abnormal
# 0 = normal
cyt_statuses <- rbinom(n_sim, 1, prop_abnormal)

# Count how many people have abnormal cells
n_abnormal <- sum(cyt_statuses == 1)
n_normal <- sum(cyt_statuses == 0)

# For the people with abnormal cells, simulate HPV
# presence based on the proportion of people with HPV
# given abnormal cells calculated in ex1.
hpv_statuses_abnormal <- rbinom(
  n_abnormal, 1,
  prop_hpv_given_abnormal
)
# For the people with normal cells, simulate HPV
# presence based on the probability of HPV
# given normal cells calculated above.
hpv_statuses_normal <- rbinom(
  n_normal, 1,
  prop_hpv_given_normal
)

# For the people with abnormal cells, simulate cervical
# cancer with a prevalence of 0.015
cc_statuses_abnormal <- rbinom(
  n_abnormal, 1,
  prop_cc_given_abnormal
)
# How many people have HPV
n_hpv <- sum(hpv_statuses_abnormal == 1) + sum(hpv_statuses_normal == 1)

# How many people have HPV and cervical cancer
n_hpv_cc <- sum(cc_statuses_abnormal == 1 & hpv_statuses_abnormal == 1)

# What proportion of people with HPV have cervical cancer
prop_cc <- n_hpv_cc / n_hpv

print(prop_cc)
```

Under this simplified model, we estimate that `r round(prop_cc*100, 2)`% of HPV-positive individuals aged 20–29 would be diagnosed with cervical cancer.

:::{.callout-note #note-cc}

This is a simplified model designed to explore conditional probability and should not be interpreted as a model of an individual's clinical risk.

:::
:::
:::

## Review {#sec-summary_0305}

[↑ top](#)

In this part of the practical, you have explored conditional probability using cervical screening data. You have seen that the probability of an outcome given a test result depends not only on the properties of the test or marker, but also on how common the outcome is in the population.


For this part of the practical, the main files you have edited are:

* `hpv.qmd`

These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}
