---
format: html
practical: Practical_3_Probability
doc-id: "0301"
author: "Katy Brown"
---

```{r setup_0301, include=FALSE}
library(tidyverse)
library(gt)
library(cowsay)
```

# Part 1: Loops {#sec-p2_loops}

::: {.callout-note .partmenu #parts-0301}

## Sections
- @sec-loops
- @sec-storing_results
- @sec-repeat_samples
- @sec-summary_0301

:::

::: {.callout-tip .objectives #objectives-0301}
#### Learning objectives
By the end of this part of the practical you will be able to:

- Explain the purpose of a `for` loop in programming.
- Use a `for` loop to repeat a task.
- Store results generated using a loop.
- Apply loops to process multiple datasets.

:::

In this practical, you will need two Quarto documents:

* `prac3_exercises.qmd` for basic exercises
* `squirrels.qmd` for a squirrel based simulation exercise
* `drosophila.qmd` for investigating mutation counts in Drosophila
* `distributions.qmd` for quick investigations of different probability distributions
* `hpv.qmd` for analysing human papillomavirus data

## For Loops {#sec-loops}

A loop is a way of repeating a task multiple times in a computer program. 

There are two main types of loop, a **for loop**, which runs once for every item in a vector (or other sequence), or a **while loop** which runs until a certain condition is met. 

In this part of the practical we will focus on **for loops**, which perform a task a fixed number of times. 

We usually loop over a vector, meaning we perform the task once for each item in the vector.

To repeat a task `n` times, we can use `seq_len(n)`. This creates a sequence from `1` to `n`. The loop then runs `n` times, once for each number in this sequence. If we run our loop as `for (i in seq_len(n))`, each time we pass through the loop, the value of `i` increases by one.

We refer to each pass through the loop as an **iteration** - we **iterate** over a loop.

In a loop, like in a function, we use curly brackets `{}` to show which code is part of the loop - so everything inside the brackets will be repeated every time we iterate through the loop.

For example:

```{r for_loop_n}
for (i in seq_len(10)) {
  print(i)
}
```
We can add steps to our loop to run on each iteration.

```{r for_loop_complex_n}
for (i in seq_len(10)) {
  # add 2 to i
  j <- i + 2
  # take the square root of j
  sj <- sqrt(j)
  # round sj to the nearest whole number
  rsj <- round(sj)
  # print rsj
  print(rsj)
}
```

We also don't have to actually use our variable `i` inside the loop, we can just use it to set the number of iterations.

::: {.callout-exercise #ex-loops_1}
In your `prac3_exercises.qmd` quarto document, generate two for loops:

1. A `for` loop which calculates and prints the **cube** of each number from 1 to 20.

::: {.callout-note #note-powers}

The operator `^` is used for powers in R, so e.g. 2^2 is 2 squared, 4^12 is 4 to the power of 12.

:::

2. This code snippet generates a random 5 letter character string.

`paste(sample(letters, 5, replace = TRUE), collapse="")`

Using a `for` loop, generate 5 such strings.

::: {.callout-answer collapse="true" #ex-loops_1_ans}

A `for` loop which calculates and prints the **cube** of each number from 1 to 20.

```{r loops_1_ans}
for (i in seq_len(20)) {
  print(i^3)
}
```

This code snippet generates a random 5 letter character string.

`paste(sample(letters, 5, replace = TRUE), collapse="")`

Using a `for` loop, generate 5 such strings.

```{r loops_2_ans}
for (i in seq_len(5)) {
  print(paste(sample(letters, 5, replace = TRUE), collapse = ""))
}
```

:::
:::

You can also use a for loop to do something to each item in a vector.

For example, in the below, we loop through the animals in our vector `species`. Inside the loop, the value of the variable `spp` changes on each iteration.

The cowsay package uses the `say()` function and the argument <code>by = [animal_name]{.orange-text}</code> (where [animal_name]{.orange-text} is the name of an animal) to print ascii art of animals (and their associates) saying things.

```{r for_loop_vector}
library(cowsay)
species <- c("frog", "stegosaurus", "squirrel", "ghost")

for (spp in species) {
  say(paste("I am your friend, a", spp), by = spp)
}
```

As before, you can add as many steps as you like to perform during each iteration.

Here, we check if each "species" name starts with the letter `s`.

```{r for_loop_vector_complex}
species <- c("monkey", "shark", "turkey", "snowman")

for (spp in species) {
  has_s <- startsWith(spp, "s")
  say(paste("Do I start with an s?", has_s), by = spp)
}
```

Remember that everything inside the brackets is part of the loop, everything outside the brackets is not.

For example, let's write a short story about bird migration, based on what we learned in practical 2.

The `cat()` function works like `print()`, but doesn't print each output on a new line.

```{r bird_story, results="hold"}
cat("A bird leaves west Africa.\n")

for (i in seq_len(25)) {
  cat("flap ")
}

cat("The bird arrives in Europe\n")
cat("6 months later...\n")
cat("A bird leaves Europe\n")

for (i in seq_len(100)) {
  cat("flap ")
}
cat("\nThe bird arrives in west Africa.")
```

Everything inside the brackets is repeated, everything outside of the brackets is not.

## Storing results from loops {#sec-storing_results}

We sometimes want to store the results of running our loop in a vector or data frame.

For example, here we store all of the cubed numbers from our example above in a vector called `cubed_nums`. Before we start the loop, we create an empty vector using `c()`. Then, each time we pass through the loop, we use `c()` again to append the new result to the end of the vector. Once the loop has finished, `cubed_nums` contains the results from every iteration.

```{r store_cubes_vec}
cubed_nums <- c()

for (i in seq_len(20)) {
  res <- i^3
  cubed_nums <- c(cubed_nums, res)
}

print(cubed_nums)
```

We can also store the results of each iteration of a loop in a data frame. One way to do this is to create an empty data frame before starting the loop using `tibble()`. During each iteration, we create a new one-row data frame containing the results from that iteration, then add it to the results data frame using `bind_rows()`.

Each time the loop runs, current_row is replaced with the results for the current species, while species_df grows by one row.

For example, we can adapt our animal loop above to store both the species name and whether it starts with "s":

```{r bind_rows}
species <- c("monkey", "shark", "turkey", "snowman")
species_df <- tibble()

for (spp in species) {
  has_s <- startsWith(spp, "s")
  
  
  current_row <- tibble(Species=spp,
                        Starts_with_S = has_s)
  species_df <- species_df |> bind_rows(current_row)
}
gt(species_df)
```


## Review {#sec-summary_0301}

[↑ top](#)

::: {.callout-note #note-summary_0301}

### Programming summary

* A `for` loop repeats a task a fixed number of times.
* Each cycle through a loop is called an **iteration**.
* You usually use a vector to determine the iterations of the loop.
* You can store the results of your loop by adding to a vector in each iteration.
* You can also store the results by adding a new row to a data frame in each iteration using `bind_rows()`.
:::

For this part of the practical, the main files you have edited are:

* `prac3_exercises.qmd`

These are the files you should select when making your Git commit.

{{< include ../includes/git_commit.qmd >}}
