---
title: "Probability_Question_Sheets"
---

## Question 1

In a natural habitat, grey wolves, foxes and coyotes co-exist. The current population
is estimated to be 60 foxes, 30 coyotes and 10 wolves. If each animal is equally likely
to be sighted first, find the probability of:
a. a coyote being sighted first
b. a wolf being sighted first
c. a fox being sighted second after either a coyote or a wolf has been sighted and
then removed from the habitat, so they can be fitted with a tracking device
(Consider: what is the impact of the first sighting on what you see next?)


```{r q1}
animals <- c(
  rep("fox", 60),
  rep("coyote", 30),
  rep("wolf", 10)
)

samps <- sample(x=animals,
                size=1000,
                replace=TRUE)
```

### a
Probability that a coyote is sighted first.

```{r q1a}
n_coy <- sum(samps == 'coyote')
print (n_coy / length(samps))
```
### b
Probability that a wolf is sighted first.

```{r q1b}
n_wolves <- sum(samps == 'wolf')
print (n_wolves / length(samps))
```
### c
Probability of a fox being sighted second after either a coyote or a wolf has been sighted and then removed from the habitat, so they can be fitted with a tracking device.

```{r q1c}
animals <- c(
  rep("fox", 60),
  rep("coyote", 30),
  rep("wolf", 10)
)

times_w_c_first = 0
times_f_second = 0
for (i in seq_len(10000)){
  samps <- sample(x=animals,
                  size=2,
                  replace=FALSE)

  if ((samps[1] == 'wolf') || (samps[1] == 'coyote')){
    times_w_c_first = times_w_c_first + 1
    if (samps[2] == 'fox'){
      times_f_second = times_f_second + 1
    }
  }
}

print (times_f_second / times_w_c_first)
```

## Question 2
A transmitter emits the signals “0” or “1” in the ratio 1:2. Two independent receivers,
labelled R1 and R2 are set up. If, unknown to the users, receiver R1 is incorrect in 2%
of the cases and receiver R2 is incorrect in 3% of the cases, find

a. the probability that R1 reports that the transmitted signal is “0”
b. the probability that R2 reports that the transmitted signal is “1”
c. following an incoming signal there is some confusion, because R1 reports
signal “0” and R2 reports signal “1”. Which of the two receivers is most likely
to be correct?

```{r q2}
transmitter_probs <- c(1/3, 2/3)
transmitter_poss_outcomes <- c(0, 1)

transmitter <- sample(x = transmitter_poss_outcomes,
                      size = 10000,
                      replace = TRUE,
                      prob=transmitter_probs)
R1_probs <- c(0.98, 0.02)
R2_probs <- c(0.97, 0.03)

R1_0 <- 0
R2_1 <- 0

R1_correct <- 0
R2_correct <- 0

n_incidents_c <- 0
for (correct in transmitter){
  incorrect = abs(correct - 1)
  
  R1_report <- sample(x = c(correct, incorrect),
                      size = 1,
                      replace = TRUE,
                      prob = R1_probs)
  R2_report <- sample(x = c(correct, incorrect),
                      size = 1,
                      replace = TRUE,
                      prob = R2_probs)
  if (R1_report == 0){
    R1_0 = R1_0 + 1
  }
  if (R2_report == 1){
    R2_1 = R2_1 + 1
  }
  if ((R1_report == 0 && R2_report == 1)){
    if (R1_report == correct){
      R1_correct = R1_correct + 1
    }
    else {
      R2_correct = R2_correct + 1
    }
  }
}
```

### a
```{r q2a}
print (R1_0 / 10000)
```

### b
```{r q2b}
print (R2_1/ 10000)
```

### c
c. following an incoming signal there is some confusion, because R1 reports
signal “0” and R2 reports signal “1”. Which of the two receivers is most likely
to be correct?

```{r q2c}
print (R1_correct / (R1_correct + R2_correct))
print (R2_correct / (R1_correct + R2_correct))
```
R2 is more likely to be correct.


