---
format: html
---
```{r include_libs, include=FALSE}
library(gt)
library(tidyverse)
combined_bird_piv <- read_csv("output/combined_birds.csv")
```

# Part 3: Visualisation {#sec-p3_visualisation}

::: {.callout-note .partmenu #parts-0203}
## Sections
- @sec-plot_recap
- @sec-blank_plot
- @sec-add_points
- @sec-axis_limits
- @sec-group_points
- @sec-add_paths
- @sec-summary_0203
:::

:::{.callout-tip .objectives #objectives-0203}
#### Learning objectives
By the end of this part of the practical you will be able to:

* Generate a new blank plot and add axes.
* Add points to a scatter plot.
* Calculate appropriate axis limits and add these to the plot.
* Change the colours and shapes of the points based on categorical data.
* Add paths connecting points based on categorical data.

:::

## Recap {#sec-plot_recap}

[↑ top](#)

Now, we want to reproduce Figure 2B from the [source publication](https://doi.org/10.1126/science.ads0532). 

Look again at the plot. You will be able to see that our data frame, combined_bird_piv, contains all the information we need.

* Breeding site
* Longitude and latitude of breeding site
* Longitude and latitude of post-Saharan stopover
* Longitude and latitude of winter non-breeding site

```{r show_bird_df}
gt(combined_bird_piv)
```

## Initiating a new plot {#sec-blank_plot}

[↑ top](#)

We will visualise the migration paths of the birds using the R plotting package `ggplot`. 

We can do this in the same Quarto document used for Part 1.

`ggplot` builds figures by starting with a blank plot and then adding layers. We first tell `ggplot` which data frame to use, then specify the variables to plot, and finally add layers such as points, lines and axes.

As we saw in Practical 1 (@sec-plots), we first need to generate a new blank plot.

We do this using the function `ggplot()`, with our data frame as an argument.

```{r blank_ggplot_birds}
ggplot(combined_bird_piv)
```

Based on the published plot, we know that our `x` axis should be **Longitude** and our `y` axis should be **Latitude**.

::: {.callout-exercise #ex-add_axes}

Look back at the plotting code from Practical 1 (@sec-plots).

Add the x and y axis to your blank plot.

::: {.callout-answer collapse="true" #ex-add_axes_ans}

```{r axes_ggplot_birds}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude)
)
```

:::
:::

## Adding points {#sec-add_points}

[↑ top](#)

A plot of longitude against latitude is simply a scatter plot - longitude determines the position on the x axis and latitude determines the position on the y axis.

To demonstrate the geom_point function, let's briefly revisit our fruit data frame from the previous practical.

```{r fruit_scatter_data}
my_data_frame <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(my_data_frame)
```

To plot this data as a scatter plot, with length on the x axis and weight on the y axis, we would do this:

```{r fruit_scatter}
ggplot(my_data_frame, aes(x = length, y = weight)) +
  geom_point()
```

::: {.callout-exercise #ex-add_points}

Going back to your bird data, plot all of the longitude and latitude points as a scatter plot.

::: {.callout-answer collapse="true" #ex-add_points_ans}

```{r points_ggplot_birds}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude)
) +
  geom_point()
```
:::
:::

If you compare this with the plot in the paper, you can see that the points are at the same positions. 

Now, we can add some additional information to make the plot more informative.

ggplot has a convenient function `annotation_borders()` to add geographical country borders to a scatter plot representing a map. We're also going to add a second function used with map data, `coord_quickmap()` - this keeps the shape of the plot consistent so the map isn't skewed.

We can add this to our scatter plot as follows. The argument `linewidth` sets the width of the lines in the map borders.

```{r map_ggplot_birds}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude)
) +
  geom_point() +
  annotation_borders(linewidth=0.2) +
  coord_quickmap()
```
## Setting axis limits {#sec-axis_limits}

[↑ top](#)

It would be clearer to zoom in just on the region containing the points, we can do this by setting the **axis limits**.

It would be helpful to know the minimum and maximum longitude and latitude of the points, we can do this with the `min()` and `max()` functions and the function `pull()`, which pulls a single column from a data frame.

e.g. to find the minimum longitude in our data frame
```{r min_long}
min_long <- combined_bird_piv |> pull(Longitude) |> min()

print (min_long)
```

::: {.callout-exercise #ex-calc_lims}
Calculate the maximum longitude and the minimum and maximum latitude, and store these as variables `max_long`, `min_lat` and `max_lat`.

::: {.callout-answer collapse="true" #ex-calc_lims_ans}
```{r axis_limits}
min_long <- combined_bird_piv |> pull(Longitude) |> min()
max_long <- combined_bird_piv |> pull(Longitude) |> max()
min_lat <- combined_bird_piv |> pull(Latitude) |> min()
max_lat <- combined_bird_piv |> pull(Latitude) |> max()
```

```{r max_long}
print (max_long)
```

```{r min_lat}
print (min_lat)
```

```{r max_lat}
print (max_lat)
```

:::
:::

We can use these numbers directly as our axis limits - but let's subtract `2` from each minimum and add `2` to each maximum so our points are not right on the edge of the plot.

We can set the x axis limits as an argument to the `coord_quickmap()` function as follows:

```{r xlim_birds}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude)
) +
  geom_point() +
  annotation_borders(linewidth=0.2) +
  coord_quickmap(xlim=c(min_long-2, max_long+2))
```
::: {.callout-exercise #ex-set_lims}
Set the y axis limits using your `min_lat` and `max_lat` variables as a second argument to `coord_quickmap()`.

::: {.callout-answer collapse="true" #ex-set_lims_ans}
```{r ylim_birds}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude)
) +
  geom_point() +
  annotation_borders(linewidth=0.2) +
  coord_quickmap(
  xlim=c(min_long-2, max_long+2),
  ylim=c(min_lat-2, max_lat+2))
```
:::
:::

## Group points {#sec-group_points}

[↑ top](#)

The aes() function controls how variables in the data are mapped to visual properties of the plot. So far we have mapped Longitude to the x-axis and Latitude to the y-axis. We can also map variables to colour, shape and many other visual properties.

Let's revisit our data frame.

```{r revisit_data_frame}
gt(combined_bird_piv)
```

We want to set the colour by the `Site` column and the shape by the `Location_Type` column.

To set colour by `Site`, we add `color = Site` as an argument to the `aes()` function.

```{r colour_site}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude, color = Site)
) +
  geom_point() +
  annotation_borders(linewidth=0.2) +
  coord_quickmap(
  xlim=c(min_long-2, max_long+2),
  ylim=c(min_lat-2, max_lat+2))
```

Setting shape by `Location_Type` is very similar, we add `shape = Location_Type` as an argument to the `aes()` function.

::: {.callout-exercise #ex-set_shape_colour}
Add this argument to the `aes()` function, along with `color=Site`, to set both the shape and the colour of the points.

::: {.callout-answer collapse="true" #ex-set_shape_colour_ans}
```{r shape_location}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)) +
  geom_point() +
  annotation_borders(linewidth=0.2) +
  coord_quickmap(
  xlim=c(min_long-2, max_long+2),
  ylim=c(min_lat-2, max_lat+2))
```
:::
:::


## Add paths {#sec-add_paths}

[↑ top](#)

Finally, we want to add paths to join the points from the same location.

To do this, we use the `geom_path()` function, grouping the points by breeding site, by adding the argument `aes(group = Site)`:

```{r connect_points}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)) +
  geom_point() +
  annotation_borders(linewidth=0.2) +
  coord_quickmap(
  xlim=c(min_long-2, max_long+2),
  ylim=c(min_lat-2, max_lat+2)) + 
  geom_path(aes(group = Site))
```

Now we have our plot! In the next part of the practical, you will generate a similar plot independently.

Meanwhile, there are various tweaks available if you want to change the appearance of your plot.

For example:

* To change the size of the points, add e.g. `size = 3` as an argument to the `geom_point()` function.
* To change the colour theme of all the data points, add `scale_color_brewer(palette = "PALETTE")`, changing PALETTE to anything on the chart on [this page](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html).
* You can change the overall theme of your plot using one of the theme options on [this page](https://ggplot2.tidyverse.org/reference/ggtheme.html)

It's worthwhile to search online for other aesthetic changes you can make to your plot, many options are available.

E.g.

```{r tweak}
ggplot(
  combined_bird_piv,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)) +
  geom_point(size=4) +
  annotation_borders(linewidth=0.2) +
  coord_quickmap(
  xlim=c(min_long-2, max_long+2),
  ylim=c(min_lat-2, max_lat+2)) + 
  geom_path(aes(group = Site)) +
  scale_color_brewer(palette='Dark2') +
  theme_bw()
```
## Review {#sec-summary_0203}

[↑ top](#)

::: {.callout-note #note-summary_0203}

### Summary

- The function `ggplot()` is used to create a new plot.
- The `aes()` function controls how variables in the data are mapped to visual properties of the plot.
- `geom_point()` is used to add points.
- `geom_path()` is used to connect related observations.
- Colour and shape can be used to display categorical variables.

:::

Remember to save your work before moving on to Part 4.

{{< include ../includes/git_commit.qmd >}}
