---
format: html
practical: Practical_2_Data_Frames_and_Plots
doc-id: "0203"
author: "Katy Brown"
---
```{r setup0203, include=FALSE}
library(gt)
library(tidyverse)
spring_data <- read_csv("output/birds_spring.csv")
```

# Part 3: Visualisation {#sec-p3_visualisation}

::: {.callout-note .partmenu #parts-0203}
## Sections
- @sec-plot_recap
- @sec-blank_plot
- @sec-add_points
- @sec-axis_limits
- @sec-group_points
- @sec-add_paths
- @sec-summary_0203
:::

:::{.callout-tip .objectives #objectives-0203}
#### Learning objectives
By the end of this part of the practical you will be able to:

* Generate a new blank plot and add axes.
* Add points to a scatter plot.
* Calculate appropriate axis limits and add these to the plot.
* Change the colours and shapes of the points based on categorical data.
* Add paths connecting points based on categorical data.

:::

## Recap {#sec-plot_recap}

[↑ top](#)

First, let's reproduce Figure 2B from the [source publication](https://doi.org/10.1126/science.ads0532). This plot shows the spring migration, which you have stored in your `spring_data` data frame.

Look again at the plot. You will be able to see that our data frame, spring_data, contains all the information we need.

* Breeding site
* Longitude and latitude of breeding site
* Longitude and latitude of post-Saharan stopover
* Longitude and latitude of winter non-breeding site

```{r show_bird_df}
gt(spring_data)
```

## Initiating a new plot {#sec-blank_plot}

[↑ top](#)

We will visualise the spring migration paths of the birds using the R plotting package `ggplot`. 

We can do this in the same Quarto document used for Part 1.

`ggplot` builds figures by starting with a blank plot and then adding layers. We first tell `ggplot` which data frame to use, then specify the variables to plot, and finally add layers such as points, lines and axes.

As we saw in Practical 1 (@sec-plots), we first need to generate a new blank plot.

We do this using the function `ggplot()`, with our data frame as an argument.

```{r blank_ggplot_birds}
ggplot(spring_data)
```

Based on the published plot, we know that our `x` axis should be **Longitude** and our `y` axis should be **Latitude**.

::: {.callout-exercise #ex-add_axes}

{{< fa solid anchor >}}

Look back at the plotting code from Practical 1 (@sec-plots).

In your **worked example** Quarto document, generate a new blank plot then add the x and y axis.

::: {.callout-answer collapse="true" #ex-add_axes_ans}

```{r add_axes_ans}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude)
)
```

:::
:::

## Adding points {#sec-add_points}

[↑ top](#)

A plot of longitude against latitude is simply a scatter plot - longitude determines the position on the x axis and latitude determines the position on the y axis.

We can add points to our plot using the `geom_point()` function, just like we used `geom_line()` to add a line in the previous practical (@sec-plots).

To demonstrate the `geom_point()` function, let's briefly revisit our fruit data frame from the previous practical.

```{r fruit_scatter_data}
my_data_frame <- tibble(
  fruit = c("apple", "pear", "banana"),
  length = c(22, 30, 55),
  weight = c(150, 165, 170)
)

gt(my_data_frame)
```

To plot this data as a scatter plot, with length on the x axis and weight on the y axis, we would do this:

```{r fruit_scatter}
ggplot(my_data_frame, aes(x = length, y = weight)) +
  geom_point()
```

::: {.callout-exercise #ex-add_points}

{{< fa solid anchor >}}

Going back to your **worked example** Quarto document, add a layer to your existing plot to plot all of the longitude and latitude points as a scatter plot.

::: {.callout-answer collapse="true" #ex-add_points_ans}

```{r add_points_ans}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude)
) +
  geom_point()
```
:::
:::

If you compare this with Fig 1B in the paper, you can see that the points are at the same positions. 

Now, we can add some additional information to make the plot more informative.

### Adding map features

ggplot has a convenient function `annotation_borders()` to add geographical country borders to a scatter plot representing a map. We're also going to add a second function used with map data, `coord_quickmap()` - this keeps the shape of the plot consistent so the map isn't skewed.

We can add this layer to our scatter plot as follows. The argument `linewidth` sets the width of the lines in the map borders.

::: {.callout-important #note-scatterplots}

The majority of the steps in this practical are useful for any scatter plot, however, this specific code chunk is specific to plotting maps.

:::

::: {.callout-note .copycode #copy-pivot}
### Include this code

In your **worked example** Quarto document, add this layer (starting at `annotation_borders`) to your existing scatterplot to add map borders.

```{r map_ggplot_birds}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap()
```
:::

## Setting axis limits {#sec-axis_limits}

[↑ top](#)

It would be clearer to zoom in just on the region containing the points, we can do this by setting the **axis limits**.

It would be helpful to know the minimum and maximum longitude and latitude of the points, we can do this with the `min()` and `max()` functions and the function `pull()`, which pulls a single column from a data frame.

e.g. to find the minimum longitude in our data frame
```{r min_long}
min_long <- spring_data |>
  pull(Longitude) |>
  min()

print(min_long)
```

::: {.callout-exercise #ex-calc_lims}

{{< fa solid anchor >}}

In your **worked example** Quarto document, before your plotting code, calculate the minimum longitude, maximum longitude and the minimum and maximum latitude, and store these as variables `min_lon`, `max_lon`, `min_lat` and `max_lat`.

::: {.callout-answer collapse="true" #ex-calc_lims_ans}
```{r calc_lims_ans1}
min_lon <- spring_data |>
  pull(Longitude) |>
  min()
max_lon <- spring_data |>
  pull(Longitude) |>
  max()
min_lat <- spring_data |>
  pull(Latitude) |>
  min()
max_lat <- spring_data |>
  pull(Latitude) |>
  max()
```

```{r calc_lims_ans2}
print(min_lon)
```

```{r calc_lims_ans3}
print(max_lon)
```

```{r calc_lims_ans4}
print(min_lat)
```

```{r calc_lims_ans5}
print(max_lat)
```

:::
:::

We can use these numbers directly as our axis limits - but let's subtract `2` from each minimum and add `2` to each maximum so our points are not right on the edge of the plot.

We can set the x axis limits as an argument to the `coord_quickmap()` function as follows:

::: {.callout-important #note-scatterplots}

For non-map data, we would most commonly use `coord_cartesian()` instead of `coord_quickmap()` here, which works in the same way.

:::

```{r xlim_birds}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(xlim = c(min_lon - 2, max_lon + 2))
```
::: {.callout-exercise #ex-set_lims}

{{< fa solid anchor >}}

Add x and y axis limits to the plot in your **worked example** Quarto document, by adding your `min_lat` and `max_lat` and `min_lon` and `max_lon` variables as arguments to `coord_quickmap()`.



::: {.callout-answer collapse="true" #ex-set_lims_ans}
```{r set_lims_ans}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(
    xlim = c(min_lon - 2, max_lon + 2),
    ylim = c(min_lat - 2, max_lat + 2)
  )
```
:::
:::

## Grouping points {#sec-group_points}

[↑ top](#)

The aes() function controls how variables in the data are mapped to visual properties of the plot. So far we have mapped Longitude to the x-axis and Latitude to the y-axis. We can also map variables to colour, shape and many other visual properties.

Let's revisit our data frame.

```{r revisit_data_frame}
gt(spring_data)
```

We want to set the colour by the `Site` column and the shape by the `Location_Type` column.

To set colour by `Site`, we add `color = Site` as an argument to the `aes()` function.

```{r colour_site}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude, color = Site)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(
    xlim = c(min_lon - 2, max_lon + 2),
    ylim = c(min_lat - 2, max_lat + 2)
  )
```

Setting shape by `Location_Type` is very similar, we add `shape = Location_Type` as an argument to the `aes()` function.

::: {.callout-exercise #ex-set_shape_colour}

{{< fa solid anchor >}}

In the plotting code in your **worked example** Quarto document, add the `shape = Location_Type` and `color = Site` arguments to the `aes()` function, to set both the shape and the colour of the points.

::: {.callout-answer collapse="true" #ex-set_shape_colour_ans}
```{r set_shape_colour_ans}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(
    xlim = c(min_lon - 2, max_lon + 2),
    ylim = c(min_lat - 2, max_lat + 2)
  )
```
:::
:::


## Adding paths {#sec-add_paths}

[↑ top](#)

Finally, we want to add paths to join the points from the same location.

To do this, we use the `geom_path()` function, grouping the points by breeding site, by adding the argument `aes(group = Site)`:

```{r connect_points}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(
    xlim = c(min_lon - 2, max_lon + 2),
    ylim = c(min_lat - 2, max_lat + 2)
  ) +
  geom_path(aes(group = Site))
```

::: {.callout-exercise #ex-add_paths}

{{< fa solid anchor >}}

In the plotting code in your **worked example** Quarto document, add the `geom_path()` function with the argument `aes(group = Site)` to your plotting code.

::: {.callout-answer collapse="true" #ex-add_paths_ans}
```{r add_paths_ans}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)
) +
  geom_point() +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(
    xlim = c(min_lon - 2, max_lon + 2),
    ylim = c(min_lat - 2, max_lat + 2)
  ) +
  geom_path(aes(group = Site))
```
:::
:::


There are various tweaks available if you want to change the appearance of your plot.

For example:

* To change the size of the points, add e.g. `size = 3` as an argument to the `geom_point()` function.
* To change the width of the line, add e.g. `linewidth = 0.5` as an argument to the `geom_line()` function.
* To change the colour theme of all the data points, add an additional layer with <code>scale_color_brewer(palette = "[PALETTE]{.orange-text}") </code>, changing [PALETTE]{.orange-text} to anything on the chart on [this page](https://r-graph-gallery.com/38-rcolorbrewers-palettes.html).
* You can change the overall theme of your plot using one of the theme options on [this page](https://ggplot2.tidyverse.org/reference/ggtheme.html)
* You can add a title by adding a layer `ggtitle()`, with the title as the only argument.

It's worthwhile to search online for other aesthetic changes you can make to your plot, many options are available.

Add any of these that you like to the plotting code in your **worked example** Quarto document.

E.g.

```{r tweak}
ggplot(
  spring_data,
  aes(x = Longitude, y = Latitude, color = Site, shape = Location_Type)
) +
  geom_point(size = 4) +
  annotation_borders(linewidth = 0.2) +
  coord_quickmap(
    xlim = c(min_lon - 2, max_lon + 2),
    ylim = c(min_lat - 2, max_lat + 2)
  ) +
  geom_path(aes(group = Site), size = 1) +
  scale_color_brewer(palette = "Dark2") +
  theme_bw() +
  ggtitle("Spring Migration Routes")
```

Now we have our plot! Compare it to Figure 1B in the publication.

::: {.callout-exercise #ex-plot_winter}

{{< fa solid flask >}}

In your **worked example** Quarto document, repeat all of the steps above with your `autumn_data` data frame, to generate your own version of Figure 1A. 

:::


## Review {#sec-summary_0203}

[↑ top](#)
You have now successfully generated two publication-quality plots based on real biological data!

::: {.callout-note #note-summary_0203}

### Summary

- The function `ggplot()` is used to create a new plot.
- The `aes()` function controls how variables in the data are mapped to visual properties of the plot.
- `geom_point()` is used to add points.
- `geom_path()` is used to connect related observations.
- Colour and shape can be used to display categorical variables.

:::

For this part of the practical, the main files you have edited are:

- `worked_example.qmd`

These are the files you should select when making your Git commit.


Remember to save and commit your work before moving on to Part 4.

{{< include ../includes/git_commit.qmd >}}
